---
title: "Mini-Project #04: Monte Carlo-Informed Selection of CUNY Retirement Plans"
author: "Juan Camilo Martinez Benavides"
date: today
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    theme: darkly
    embed-resources: true
execute:
  warning: false
  message: false
---

## Executive Summary

Selecting a retirement plan is a critical financial decision for CUNY faculty. This project evaluates the Teachers Retirement System (TRS) and the Optional Retirement Plan (ORP) using historical economic data and Monte Carlo simulations. By modeling contributions, investment returns, and inflation-adjusted benefits, we provide a data-driven comparison to assist in this irrevocable decision.

## Introduction

Selecting a retirement plan is one of the most critical financial decisions for new CUNY faculty, as it has permanent implications for their long-term financial security. Once chosen, the plan cannot be changed, making it essential to consider the potential risks and rewards.

## Objectives

The primary goal of this project is to equip new CUNY faculty with an evidence-based recommendation for selecting the optimal retirement plan (TRS or ORP). This involves analyzing historical economic and financial data to assess how each plan performs under different economic conditions. We will utilize the following datasets:

-   Rate of Inflation: Inflation erodes the purchasing power of retirement benefits. For TRS, it determines the annual adjustment in benefits, while for ORP, it impacts the real value of savings and withdrawals.
-   Rate of Wage Growth: Wage growth directly influences contributions to both plans. For TRS, it affects the final average salary (FAS), which determines retirement payouts. For ORP, higher wages lead to larger investment contributions.
-   US Equity Market Returns: Equity market performance is crucial for ORP participants, as their investments grow with market returns. US equities represent a significant portion of most retirement portfolios.
-   International Equity Market Returns: Diversification through international equities reduces risk and enhances returns for ORP participants. It provides a broader perspective on investment performance beyond the US market.
-   Bond Returns: Bonds are a critical component of retirement portfolios, especially for ORP participants as they approach retirement. Bonds provide stability and reduce volatility in investment returns.
-   Short-Term Debt Returns: Short-term debt, such as Treasury yields, represents a safe investment option with lower risk. It is particularly relevant for retirees seeking liquidity and capital preservation.

By combining insights from these datasets, we will simulate potential retirement outcomes for each plan, identify risks such as running out of funds, and quantify the financial benefits of each option. This comprehensive approach ensures that our recommendation aligns with the diverse financial priorities of faculty members.

### Libraries

The following libraries are used throughout the project:

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
library(httr2)       # API requests
library(dplyr)       # Data manipulation
library(ggplot2)     # Visualization
library(zoo)         # Time-series data
library(DT)          # Interactive tables
library(ggcorrplot)  # Correlation plots
library(lubridate)   # Date/time handling
library(gt)          # Publication-quality tables
library(plotly)      # Interactive plots
library(tidyr)       # Data reshaping




```

## Data Collection

For this project, we will use data from two economic and financial data sources:

1.  **AlphaVantage**: A commercial stock market data provider.
2.  **FRED**: The Federal Reserve Economic Data repository maintained by the Federal Reserve Bank of St. Louis.

FRED is free to access, but AlphaVantage is a commercial service requiring a subscription. For this mini-project, the free tier of AlphaVantage will suffice.

### Links to Resources

-   [AlphaVantage](https://www.alphavantage.co)
-   [FRED (Federal Reserve Economic Data)](https://fred.stlouisfed.org)

## Data Acquisition

We collect data on a monthly basis for the last 15 years using the following APIs:

To do so, we use their APIs:

```{r}
# Define the base URL for the FRED API
fred_url <- "https://api.stlouisfed.org/fred/series/observations"

# Define the base URL for the AlphaVantage API
alphavantage_url <- "https://www.alphavantage.co/query"

# Read FRED API key from a file
fred_key <- readLines("fred_key.txt") # Ensure this file contains only your FRED API key

# Read AlphaVantage API key from a file
alphavantage_key <- readLines("alphavantage_key.txt") # Ensure this file contains only your AlphaVantage API key

# Montly basis for the last 15 years
Time_Montly =as.Date("2009-01-01")
```

## Fetching Data

Next, we fetch the data:

- 1. Fetching Rate of Inflation (CPI for NYC Metro Area) from FRED

```{r}
#| echo: true
#| message: false
#| warning: false
# Fetch Inflation Data from FRED (CPI for NYC Metro Area)
inflation_req <- request(fred_url) |>
  req_url_query(
    series_id = "CUUR0400SA0",  # CPI for All Urban Consumers: NYC Metro Area
    api_key = fred_key,
    file_type = "json"
  )

inflation_resp <- inflation_req |> req_perform()
inflation_data <- resp_body_json(inflation_resp)$observations

# Convert to DataFrame
inflation_df <- data.frame(
  date = as.Date(sapply(inflation_data, function(x) x$date)),
  inflation_rate = as.numeric(sapply(inflation_data, function(x) x$value))
)

# Filter the data for the last 15 years
inflation_df <- inflation_df |> filter(date >= Time_Montly)
```

- 2. Fetching Rate of Wage Growth from FRED, We’ll use the AHETPI series to measure wage growth.

```{r}
# Fetch Wage Growth Data from FRED
wage_growth_req <- request(fred_url) |>
  req_url_query(
    series_id = "AHETPI", # Average Hourly Earnings of Production and Nonsupervisory Employees
    api_key = fred_key,
    file_type = "json"
  )

wage_growth_resp <- wage_growth_req |> req_perform()
wage_growth_data <- resp_body_json(wage_growth_resp)$observations

# Convert to a DataFrame
wage_growth_df <- data.frame(
  date = as.Date(sapply(wage_growth_data, function(x) x$date)), # Parse date
  wage_growth = as.numeric(sapply(wage_growth_data, function(x) x$value)) # Convert to numeric
)

# Filter data for the last 15–20 years
wage_growth_df <- wage_growth_df |> filter(date >= Time_Montly)
```

- 3. Fetching US Equity Market Returns from AlphaVantage

```{r}
# Fetch US Equity Market Data from AlphaVantage
us_equity_req <- request(alphavantage_url) |>
  req_url_query(
    `function` = "TIME_SERIES_MONTHLY_ADJUSTED",
    symbol = "SPY", # S&P 500 ETF
    apikey = alphavantage_key
  )
us_equity_resp <- us_equity_req |> req_perform()
us_equity_data <- resp_body_json(us_equity_resp)$`Monthly Adjusted Time Series`

# Convert to DataFrame
us_equity_df <- data.frame(
  date = as.Date(names(us_equity_data)), # Extract dates
  us_equity_close = as.numeric(sapply(us_equity_data, function(x) x$`5. adjusted close`)) # Extract adjusted close prices
)

# Standardize dates to the first of each month
us_equity_df$date <- as.Date(format(us_equity_df$date, "%Y-%m-01"))

# Filter for the past 15 years
us_equity_df <- us_equity_df |> filter(date >= Time_Montly)
```

- 4. Fetching International Equity Market Returns from AlphaVantage

```{r}
# Fetch International Equity Market Data from AlphaVantage
intl_req <- request(alphavantage_url) |>
  req_url_query(
    `function` = "TIME_SERIES_MONTHLY_ADJUSTED",
    symbol = "VEA", # Vanguard FTSE Developed Markets ETF
    apikey = alphavantage_key
  )

intl_resp <- intl_req |> req_perform()
intl_data <- resp_body_json(intl_resp)$`Monthly Adjusted Time Series`

# Convert the fetched data into a data frame
intl_equity_df <- data.frame(
  date = as.Date(names(intl_data)),
  intl_equity_close = as.numeric(sapply(intl_data, function(x) x$`5. adjusted close`))
)

# Standardize dates to the first of each month
intl_equity_df$date <- as.Date(format(intl_equity_df$date, "%Y-%m-01"))

# Filter for the past 15 years
intl_equity_df <- intl_equity_df |> filter(date >= Time_Montly)
```

- 5. Fetching Bond Returns from AlphaVantage

```{r}
#| echo: true
#| message: false
#| warning: false
# Fetch Bond Yield Data (Daily)
bond_req <- request(fred_url) |>
  req_url_query(
    series_id = "DGS10", # 10-Year Treasury Yield
    api_key = fred_key,
    file_type = "json"
  )

bond_resp <- bond_req |> req_perform()
bond_data <- resp_body_json(bond_resp)$observations

# Convert to DataFrame
bond_df <- data.frame(
  date = as.Date(sapply(bond_data, function(x) x$date)),
  bond_yield = as.numeric(sapply(bond_data, function(x) x$value))
)

# Downsample to Monthly Averages
bond_df <- bond_df |>
  mutate(month = format(date, "%Y-%m")) |># Extract year-month for grouping
  group_by(month) |>
  summarize(
    bond_yield = mean(bond_yield, na.rm = TRUE), # Average yield for the month
    .groups = "drop"
  ) |>
  mutate(date = as.Date(paste0(month, "-01"))) |># Convert back to Date
  select(date, bond_yield)

# Filter for the last 15 years
bond_df <- bond_df |> filter(date >= Time_Montly)
```

- 6.  Fetching Short-Term Debt Returns from AlphaVantage: for this dataset, we need to downsample the daily data to monthly averages to align with the other datasets.

```{r}
#| echo: true
#| message: false
#| warning: false
# Fetch Short-Term Debt Data (2-Year Treasury Yield) - Attempt Monthly Query
short_term_req <- request(fred_url) |>
  req_url_query(
    series_id = "DGS2", # 2-Year Treasury Yield
    frequency = "m", # Request monthly data directly
    api_key = fred_key,
    file_type = "json"
  )

short_term_resp <- short_term_req |> req_perform()
short_term_data <- resp_body_json(short_term_resp)$observations

# Convert JSON to DataFrame
short_term_df <- data.frame(
  date = as.Date(sapply(short_term_data, function(x) x$date), format = "%Y-%m-%d"),
  short_term_yield = as.numeric(sapply(short_term_data, function(x) x$value))
)

# Filter for the Last 15 Years
short_term_df <- short_term_df |> filter(date >= Time_Montly)
```

## Data Preprocessing and Cleaning

With the data collected, we proceed to preprocessing. We unify the datasets and handle missing values using bootstrapping.

```{r}
#| echo: true
#| message: false
#| warning: false

# Standardize dates to the first of each month for all datasets
datasets <- list(inflation_df, wage_growth_df, us_equity_df, intl_equity_df, bond_df, short_term_df)
datasets <- datasets |>
  lapply(\(df) {
    df$date <- as.Date(format(df$date, "%Y-%m-01"))
    df
  })

# Unpack datasets back
inflation_df <- datasets[[1]]
wage_growth_df <- datasets[[2]]
us_equity_df <- datasets[[3]]
intl_equity_df <- datasets[[4]]
bond_df <- datasets[[5]]
short_term_df <- datasets[[6]]

# Create master date sequence
date_seq <- data.frame(date = seq(from = as.Date("2009-01-01"), 
                                  to = as.Date(format(Sys.Date(), "2024-10-01")), 
                                  by = "month"))

# Perform left joins for all datasets
combined_data <- date_seq |>
  left_join(inflation_df, by = "date") |>
  left_join(wage_growth_df, by = "date") |>
  left_join(us_equity_df, by = "date") |>
  left_join(intl_equity_df, by = "date") |>
  left_join(bond_df, by = "date") |>
  left_join(short_term_df, by = "date")

# Handle missing data
combined_data <- combined_data |>
  mutate(
    inflation_rate = replace(inflation_rate,
                             is.na(inflation_rate),
                             sample(na.omit(inflation_rate), size = sum(is.na(inflation_rate)), replace = TRUE)),
    wage_growth = replace(wage_growth,
                          is.na(wage_growth),
                          sample(na.omit(wage_growth), size = sum(is.na(wage_growth)), replace = TRUE)),
    us_equity_close = replace(us_equity_close,
                              is.na(us_equity_close),
                              sample(na.omit(us_equity_close), size = sum(is.na(us_equity_close)), replace = TRUE)),
    intl_equity_close = replace(intl_equity_close,
                                is.na(intl_equity_close),
                                sample(na.omit(intl_equity_close), size = sum(is.na(intl_equity_close)), replace = TRUE)),
    bond_yield = replace(bond_yield,
                         is.na(bond_yield),
                         sample(na.omit(bond_yield), size = sum(is.na(bond_yield)), replace = TRUE)),
    short_term_yield = replace(short_term_yield,
                               is.na(short_term_yield),
                               sample(na.omit(short_term_yield), size = sum(is.na(short_term_yield)), replace = TRUE))
  ) |>
  mutate(across(where(is.numeric), ~ round(., 2)))

# Display dataset interactively
DT::datatable(combined_data, options = list(pageLength = 10, scrollX = TRUE))
```

## Initial Analysis

We visualize the data to understand trends and relationships, comparing key variables specifically:

```{r}
#| echo: true
#| message: false
#| warning: false
# Create plot for Inflation Rate and Wage Growth with corrected labels and separate y-axes
p1 <- ggplot(combined_data, aes(x = date)) +
  # Inflation Rate
  geom_line(aes(y = inflation_rate), color = "blue") +
  scale_y_continuous(
    name = "Inflation Rate Value (NYC)",  # Corrected label for inflation rate
    sec.axis = sec_axis(~ . * 0, name = " ")  # Create empty secondary axis for separation
  ) +
  # Wage Growth (scaled for visibility)
  geom_line(aes(y = wage_growth * 10), color = "green") +
  scale_y_continuous(
    name = "Inflation Rate Value (NYC)",  # Corrected label for wage growth
    sec.axis = sec_axis(~ . / 10, name = "Wage Growth (Hourly Earnings in $)")  # Corrected secondary label
  ) +
  labs(title = "Inflation Rate and Wage Growth Over Time", x = "Date") +
  theme_minimal() +
  theme(
    axis.title.y.left = element_text(color = "blue", size = 12),
    axis.title.y.right = element_text(color = "green", size = 12)
  )

# Create plot for US Equity and International Equity
p2 <- ggplot(combined_data, aes(x = date)) +
  # US Equity Return
  geom_line(aes(y = us_equity_close), color = "red") +  
  scale_y_continuous(
    name = "US Equity Return (Adjusted Close)"  
  ) +
  # International Equity Return (scaled for visibility)
  geom_line(aes(y = intl_equity_close * 10), color = "purple") +  
  scale_y_continuous(
    name = "US Equity Return (Adjusted Close)",
    sec.axis = sec_axis(~ . / 10, name = "Intl Equity Return")  
  ) +
  labs(title = "US and International Equity Returns Over Time", x = "Date") +
  theme_minimal() +
  theme(
    axis.title.y.left = element_text(color = "red", size = 12),
    axis.title.y.right = element_text(color = "purple", size = 12)
  )

# Create plot for Bond Yield and Short-Term Yield 
p3 <- ggplot(combined_data, aes(x = date)) +
  # Bond Yield
  geom_line(aes(y = bond_yield * 2), color = "orange") +  
  scale_y_continuous(
    name = "Bond Yield (%)", 
    sec.axis = sec_axis(~ . / 2, name = "Bond Yield (%)")  
  ) +
  # Short-Term Yield
  geom_line(aes(y = short_term_yield * 2), color = "brown") +  
  scale_y_continuous(
    name = "Short-Term Yield (%)", 
    sec.axis = sec_axis(~ . / 2, name = "Short-Term Yield (%)")  
  ) +
  labs(title = "Bond Yield and Short-Term Yield Over Time", x = "Date") +
  theme_minimal() +
  theme(
    axis.title.y.left = element_text(color = "orange", size = 12),
    axis.title.y.right = element_text(color = "brown", size = 12)
  )

# Display all three plots
p1
p2
p3

```

We can see a clear correlation between some of this plots. For example, the inflation rate and wage growth are highly correlated, as well as the US and International Equity Returns.

### Correlation matrix

We will calculate the correlation matrix to understand the relationships between the variables. This will help us identify any significant correlations that can inform our analysis and decision-making process.

```{r}
# Select only numeric columns
numeric_data <- combined_data |>
  select(where(is.numeric))

# Compute the correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")

# Visualize the correlation matrix using ggcorrplot
ggcorrplot(cor_matrix, 
           type = "lower",          # Only show the lower triangle
           lab = TRUE,              # Show correlation coefficient labels
           lab_size = 3,            # Adjust label size
           method = "circle",       # Use circle shapes for correlation values
           colors = c("red", "white", "blue"), # Color scale
           title = "Correlation Matrix", # Plot title
           ggtheme = theme_minimal(), # Minimalistic theme
           tl.cex = 12,             # Title label size
           tl.col = "black"         # Title text color
)
```

In this initial analysis we examine the relationships between several key economic factors, including inflation, wage growth, stock market returns (both US and international), bond yields, and short-term debt yields. Understanding how these factors are correlated can provide valuable insights for making informed financial decisions, particularly for retirement planning.

## Key Findings:

### 1. Inflation Rate and Wage Growth:

-   **Correlation**: 0.98 (Very strong)
-   **Interpretation**: When inflation increases, wage growth tends to follow suit. This suggests that as the cost of living rises, your wages are likely to increase as well, which helps you keep up with inflation.

### 2. Stock Market Returns:

#### US and International Markets:

-   **Correlation**: 0.97 (Very strong)
-   **Interpretation**: US and international stock markets have a highly positive correlation, meaning they often move in the same direction. This makes diversification between US and international markets a good strategy for reducing risk while maintaining growth potential.

#### US Equity Returns and Bond Yields:

-   **Correlation**: 0.68 (Moderate)
-   **Interpretation**: The moderate correlation suggests that when US stocks perform well, bond yields may not follow the same trend. Investors need to balance their portfolios between stocks and bonds, taking into account their risk tolerance and time horizon.

### 3. Inflation and Other Variables:

#### Inflation and US Stock Returns:

-   **Correlation**: 0.92 (Strong)
-   **Interpretation**: Inflation and US stock returns are highly correlated. As inflation rises, US stocks tend to perform well, making stocks a potential hedge against inflation.

#### Inflation and International Stock Returns:

-   **Correlation**: 0.97 (Very strong)
-   **Interpretation**: Like US stocks, international stocks also tend to rise with inflation, making global diversification a viable strategy for mitigating inflation risk.

#### Inflation and Bond Yields:

-   **Correlation**: 0.72 (Moderate)
-   **Interpretation**: Inflation and bond yields are moderately correlated. Rising inflation generally leads to higher bond yields, which can help protect fixed-income investments from losing purchasing power.

### 4. Bond Yield and Short-Term Yield:

-   **Correlation**: 0.75 (Strong)
-   **Interpretation**: Bond yields and short-term yields tend to move in the same direction, which suggests that both types of bonds respond similarly to changes in interest rates.


## Implications for Retirement Planning:

-   **Wage Growth and Inflation**: The close correlation between wage growth and inflation means that as prices rise, your income should likely increase, helping you maintain purchasing power in retirement.

-   **Equity Market Performance**: The strong correlation between US and international equity returns allows for diversification between these markets without sacrificing performance.

-   **Bonds and Short-Term Yields**: Understanding the relationship between bonds and short-term debt yields is crucial for managing fixed-income investments in a rising interest rate environment.

# Historical Comparison of TRS and ORP

We compare these two models using the same initial salary and service years to determine the optimal choice. Key assumptions include:


## Initial Setup and Parameters
- **Initial Salary**: $60,000.
- **Years of Service**: 15 years.
- **Salary Growth**: Modeled using year-over-year CPI growth, calculated from the `combined_data` dataset.

## TRS (Teacher Retirement System) Benefit Calculation
1. Calculated annual salaries over the 15-year period by applying CPI growth.
2. Determined the **Final Average Salary (FAS)** as the average of the last 3 years of salaries.
3. Computed the **Annual TRS Benefit** based on:
   - **Years of Service**:
     - 1.67% of FAS for each year of service up to 20 years.
     - Special rules applied for exactly 20 years or more than 20 years.
   - **FAS**: Multiplied by the benefit rate and years of service.
4. Displayed both the annual and monthly TRS benefits.

## Contributions (Employee and Employer)
1. Calculated annual employee and employer contributions based on salary and contribution rules:
   - **Employee Contributions**:
     - 3% to 6% of salary, determined by salary brackets.
   - **Employer Contributions**:
     - 8% for the first 7 years and 10% for subsequent years.
2. Summed contributions for each year and totaled them over the 15-year period.
3. Stored detailed data in a table for yearly contributions.

## ORP (Optional Retirement Plan) Investment Simulation
1. **Portfolio Allocation**:
   - Adjusted allocations for US equities, international equities, and bonds based on age brackets:
     - Ages 50-59: Higher equity allocation.
     - Ages 60-74: Increased bond allocation to have more stability. 
2. Calculated annual performance for each asset class using growth rates from the `combined_data` dataset:
   - US and international equity growth based on stock market performance.
   - Bond yields calculated as annual averages.
3. Simulated yearly portfolio performance using contributions and compound growth.
4. Computed the **Total Accumulated Value** at the end of 15 years.
5. Estimated the **Annual ORP Benefit** as 4% of the final accumulated value.

---
The following code snippets will calculate the retirement benefits for both TRS and ORP based on the assumptions outlined above. We will compare the monthly benefits and accumulated funds for each plan to provide a comprehensive analysis of the retirement options.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true

# Set initial salary and define years of service
initial_salary <- 60000
years_of_service <- 15  # Adjusted to 15 years

# Initialize a vector to store salaries for each year
salaries <- numeric(years_of_service)
salary_growth <- numeric(years_of_service)
cpi_growth <- numeric(years_of_service)
salary_increase <- numeric(years_of_service)

# Set the first salary as the initial salary
salaries[1] <- initial_salary

# Loop through the combined data for yearly salary growth calculation
for (i in 2:years_of_service) {
  # We need CPI data for each September to calculate the salary for the next year
  # Calculate the CPI growth between September (from the previous year) and September of the current year
  
  september_cpi_previous_year <- combined_data$inflation_rate[which(month(combined_data$date) == 9 & year(combined_data$date) == 2008 + (i - 1))]
  september_cpi_current_year <- combined_data$inflation_rate[which(month(combined_data$date) == 9 & year(combined_data$date) == 2008 + i)]
  
  # Calculate the salary growth rate based on CPI change
  cpi_growth[i] <- september_cpi_current_year / september_cpi_previous_year
  salary_growth[i] <- cpi_growth[i]
  
  # Calculate the increase in salary
  salary_increase[i] <- salaries[i-1] * (cpi_growth[i] - 1)  # Difference from the previous salary
  
  # Apply the salary growth to the previous year's salary
  salaries[i] <- salaries[i-1] * cpi_growth[i]  
}

# Calculate Final Average Salary (FAS) - which is the average of the last 3 years' salary
fas <- mean(salaries[(years_of_service-2):years_of_service])

# Calculate the Annual Retirement Benefit based on years of service
if (years_of_service <= 20) {
  annual_benefit_trs <- 0.0167 * fas * years_of_service
} else if (years_of_service == 20) {
  annual_benefit_trs <- 0.0175 * fas * years_of_service
} else {
  annual_benefit_trs <- (0.35 + 0.02 * years_of_service) * fas
}

# Prepare the data for the table
trs_benefit_data <- data.frame(
  Benefit_Type = c("Annual TRS Benefit", "Monthly TRS Benefit"),
  Amount = c(round(annual_benefit_trs, 2), round(annual_benefit_trs / 12, 2))
)

# Create and display the table using gt
trs_benefit_data |>
  gt() |>
  tab_header(
    title = "TRS Benefit Summary",
    subtitle = "Annual and Monthly TRS Benefits"
  ) |>
  fmt_currency(
    columns = "Amount",
    currency = "USD"
  ) |>
  cols_label(
    Benefit_Type = "Benefit Type",
    Amount = "Amount ($)"
  ) |>
  tab_options(
    table.font.size = "medium",
    heading.align = "center"
  )

```
The table above shows the estimated TRS benefits. Next, we calculate contributions for the ORP plan.

```{r}
#| echo: true
#| message: false
#| warning: false

# Initialize parameters
initial_salary <- 60000  # Starting salary
years_of_service <- 15   # Number of years to simulate

# Set up salary contribution brackets
salary_brackets <- c(45000, 55000, 75000, 100000)

# Initialize vectors to store results
salaries <- numeric(years_of_service)
employee_contributions <- numeric(years_of_service)
employer_contributions <- numeric(years_of_service)
total_contributions <- numeric(years_of_service)

# Initialize the first year's salary
salaries[1] <- initial_salary

# Loop through each year and compute the contributions
for (i in 1:years_of_service) {
  
  # Get the salary for the year
  if (i == 1) {
    salary <- initial_salary
  } else {
    # Calculate salary growth based on CPI (only starting from the second year)
    september_cpi_previous_year <- combined_data$inflation_rate[which(month(combined_data$date) == 9 & year(combined_data$date) == 2009 + (i - 1))]
    september_cpi_current_year <- combined_data$inflation_rate[which(month(combined_data$date) == 9 & year(combined_data$date) == 2009 + i)]
    
    # Calculate the salary growth rate based on CPI change (year over year growth)
    cpi_growth <- september_cpi_current_year / september_cpi_previous_year
    salary <- salaries[i - 1] * cpi_growth  # Update salary based on CPI growth
  }
  
  # Store the salary for each year
  salaries[i] <- salary
  
  # Employee contribution calculation based on salary brackets (annual contribution)
  if (salary <= salary_brackets[1]) {
    employee_contribution_annual <- salary * 0.03
  } else if (salary <= salary_brackets[2]) {
    employee_contribution_annual <- salary * 0.035
  } else if (salary <= salary_brackets[3]) {
    employee_contribution_annual <- salary * 0.045
  } else if (salary <= salary_brackets[4]) {
    employee_contribution_annual <- salary * 0.0575
  } else {
    employee_contribution_annual <- salary * 0.06
  }
  
  # Employer contribution: 8% for the first 7 years, 10% after (annual contribution)
  if (i <= 7) {
    employer_contribution_annual <- salary * 0.08
  } else {
    employer_contribution_annual <- salary * 0.10
  }
  
  # Total contribution for the year
  total_contribution_annual <- employee_contribution_annual + employer_contribution_annual
  
  # Store contributions
  employee_contributions[i] <- employee_contribution_annual
  employer_contributions[i] <- employer_contribution_annual
  total_contributions[i] <- total_contribution_annual
}

# Create the table with the results for each year
yearly_contributions_data <- data.frame(
  Year = 1:years_of_service,
  Salary = round(salaries, 2),
  Employee_Contribution = round(employee_contributions, 2),
  Employer_Contribution = round(employer_contributions, 2),
  Total_Contribution = round(total_contributions, 2)
)

# Calculate cumulative contributions
yearly_contributions_data$Cumulative_Employee_Contribution <- cumsum(yearly_contributions_data$Employee_Contribution)
yearly_contributions_data$Cumulative_Employer_Contribution <- cumsum(yearly_contributions_data$Employer_Contribution)
yearly_contributions_data$Cumulative_Total_Contribution <- cumsum(yearly_contributions_data$Total_Contribution)

# Visualization for cumulative contributions
library(ggplot2)

ggplot(yearly_contributions_data, aes(x = Year)) +
  geom_line(aes(y = Cumulative_Employee_Contribution, color = "Cumulative Employee Contribution"), size = 1) +
  geom_line(aes(y = Cumulative_Employer_Contribution, color = "Cumulative Employer Contribution"), size = 1) +
  geom_line(aes(y = Cumulative_Total_Contribution, color = "Cumulative Total Contribution"), size = 1) +
  labs(
    title = "Cumulative Contributions Over Time",
    x = "Year of Service",
    y = "Cumulative Contribution Amount ($)",
    color = "Legend"
  ) +
  theme_minimal()
```

```{r}
# Initialize parameters for ORP calculations
# Age 50 to 59
equity_us_alloc_50_59 <- 0.47  # US Equities allocation
equity_international_alloc_50_59 <- 0.32  # International Equities allocation
bonds_alloc_50_59 <- 0.21  # Bonds allocation

# Age 60 to 74
equity_us_alloc_60_74 <- 0.34  # US Equities allocation
equity_international_alloc_60_74 <- 0.23  # International Equities allocation
bonds_alloc_60_74 <- 0.43  # Bonds allocation

# Initialize vectors to store results
performance_us_equity <- numeric(years_of_service)
performance_international_equity <- numeric(years_of_service)
performance_bonds <- numeric(years_of_service)
total_accumulated <- numeric(years_of_service)  # Total accumulated amount

# Starting balance
total_accumulated[1] <- yearly_contributions_data$Total_Contribution[1]

# Start contributing at age 50 in 2009
starting_age <- 50

# Loop through each year to compute performance
for (i in 1:years_of_service) {
  current_year <- 2009 + i - 1  # Current year
  next_year <- current_year + 1  # Next year
  current_age <- starting_age + i - 1  # Current age
  
  # Assign allocations based on age
  if (current_age >= 50 && current_age <= 59) {
    equity_us_alloc <- equity_us_alloc_50_59
    equity_international_alloc <- equity_international_alloc_50_59
    bonds_alloc <- bonds_alloc_50_59
  } else if (current_age >= 60 && current_age <= 74) {
    equity_us_alloc <- equity_us_alloc_60_74
    equity_international_alloc <- equity_international_alloc_60_74
    bonds_alloc <- bonds_alloc_60_74
  }
  
  # Ensure we have data for the future January, or use November of the current year if not available
  if (i == years_of_service) {
    # For the last year, use November data instead of January of the next year
    us_equity_growth <- combined_data$us_equity_close[which(year(combined_data$date) == current_year & month(combined_data$date) == 11)] /
                        combined_data$us_equity_close[which(year(combined_data$date) == current_year & month(combined_data$date) == 1)]
    
    intl_equity_growth <- combined_data$intl_equity_close[which(year(combined_data$date) == current_year & month(combined_data$date) == 11)] /
                          combined_data$intl_equity_close[which(year(combined_data$date) == current_year & month(combined_data$date) == 1)]
    
    bond_yield_annual <- mean(combined_data$bond_yield[which(year(combined_data$date) == current_year & month(combined_data$date) <= 11)]) / 100 + 1
  } else {
    # For all other years, use January of the next year
    us_equity_growth <- combined_data$us_equity_close[which(year(combined_data$date) == next_year & month(combined_data$date) == 1)] /
                        combined_data$us_equity_close[which(year(combined_data$date) == current_year & month(combined_data$date) == 1)]
    
    intl_equity_growth <- combined_data$intl_equity_close[which(year(combined_data$date) == next_year & month(combined_data$date) == 1)] /
                          combined_data$intl_equity_close[which(year(combined_data$date) == current_year & month(combined_data$date) == 1)]
    
    bond_yield_annual <- mean(combined_data$bond_yield[which(year(combined_data$date) == current_year)]) / 100 + 1
  }

  # Store individual performance values
  performance_us_equity[i] <- us_equity_growth
  performance_international_equity[i] <- intl_equity_growth
  performance_bonds[i] <- bond_yield_annual  # Already adjusted with +1
  
  # Calculate total accumulated with contributions and compound effect
  if (i == 1) {
    # First year: Apply performance to the first contribution
    total_accumulated[i] <- yearly_contributions_data$Total_Contribution[i] * (
                              (equity_us_alloc * performance_us_equity[i]) +
                              (equity_international_alloc * performance_international_equity[i]) +
                              (bonds_alloc * performance_bonds[i])
                            )
  } else {
    # Subsequent years: Include previous accumulated total and current contributions
    total_accumulated[i] <- (total_accumulated[i - 1] + yearly_contributions_data$Total_Contribution[i]) * (
                              (equity_us_alloc * performance_us_equity[i]) +
                              (equity_international_alloc * performance_international_equity[i]) +
                              (bonds_alloc * performance_bonds[i])
                            )
  }
}

# Create a data frame with the results
yearly_performance_data <- data.frame(
  Year = 2009:(2009 + years_of_service - 1),
  Age = starting_age:(starting_age + years_of_service - 1),
  US_Equity_Performance = round(performance_us_equity, 4),
  International_Equity_Performance = round(performance_international_equity, 4),
  Bonds_Performance = round(performance_bonds, 4),
  Total_Accumulated = round(total_accumulated, 2)
)
library(gt)

annual_benefit_opr <- 0.04 * yearly_performance_data$Total_Accumulated[years_of_service]
monthly_benefit_opr <- annual_benefit_opr / 12

opr_benefit_table <- data.frame(
  Metric = c("Annual OPR Benefit", "Monthly OPR Benefit"),
  Amount = c(round(annual_benefit_opr, 2), round(monthly_benefit_opr, 2))
)

opr_benefit_table |>
  gt() |>
  tab_header(
    title = "OPR Benefit Summary",
    subtitle = paste("Year", years_of_service, "Calculation")
  ) |>
  fmt_currency(
    columns = "Amount",
    currency = "USD"
  ) |>
  cols_label(
    Metric = "Benefit Type",
    Amount = "Amount ($)"
  ) |>
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels(everything())
  )
```

```{r}
# Calculate total accumulated contributions over time
total_contributions <- cumsum(yearly_contributions_data$Total_Contribution)

# Create a data frame for plotting
plot_data <- data.frame(
  Year = 2009:(2009 + years_of_service - 1),
  Total_Accumulated = total_accumulated,
  Total_Contributions = total_contributions
)

# Create the interactive graph
plot <- plot_ly(plot_data, x = ~Year) |>
  add_trace(y = ~Total_Accumulated, type = 'scatter', mode = 'lines+markers', name = 'Compounding $',
            hovertemplate = 'Year: %{x}<br>Total Accumulated: %{y:.2f}<extra></extra>') |>
  add_trace(y = ~Total_Contributions, type = 'scatter', mode = 'lines+markers', name = 'Total Contributions',
            hovertemplate = 'Year: %{x}<br>Total Contributions: %{y:.2f}<extra></extra>') |>
  layout(
    title = 'Interactive Graph: Total Accumulation vs. Contributions',
    xaxis = list(title = 'Year'),
    yaxis = list(title = 'Amount ($)'),
    legend = list(orientation = 'h', x = 0.5, xanchor = 'center')
  )

plot


```
In this image we can see clearly the power of compounding. The total accumulated amount grows over time due to the contributions and the compound effect of the investment returns. This visualization provides a clear representation of how the retirement savings can grow over the years, highlighting the importance of consistent contributions and wise investment choices.


## Comparative Analysis of TRS and ORP

```{r}
# Calculate monthly benefits
monthly_benefit_opr <- annual_benefit_opr / 12
monthly_benefit_trs <- annual_benefit_trs / 12

# Create a data frame for plotting
benefit_data <- data.frame(
  Benefit_Type = c("Monthly OPR Benefit", "Monthly TRS Benefit"),
  Amount = c(monthly_benefit_opr, monthly_benefit_trs)
)

# Create the bar plot
benefit_plot <- ggplot(benefit_data, aes(x = Benefit_Type, y = Amount, fill = Benefit_Type)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0("$", round(Amount, 2))), vjust = -0.5, size = 4) +
  labs(
    title = "Comparison of Monthly OPR and TRS Benefits",
    x = "Benefit Type",
    y = "Amount ($)"
  ) +
  theme_minimal()

# Print the plot
print(benefit_plot)
```

The bar chart illustrates a comparison between the **Monthly OPR Benefit** and the **Monthly TRS Benefit**. Key insights from the visualization are as follows:

1.  **Monthly OPR Benefit**:
    -   The monthly benefit derived from the OPR system is **\$883.35**, as shown by the first bar.
    -   This represents a smaller portion compared to the TRS benefit, reflecting the structure of OPR benefit allocations.
2.  **Monthly TRS Benefit**:
    -   The monthly benefit derived from the TRS system is **\$1769.43**, as shown by the taller second bar.
    -   This indicates a significantly higher payout, suggesting that the TRS system may provide a more robust financial benefit for retirees compared to the OPR system.
3.  **Comparison**:
    -   The TRS benefit is nearly double the OPR benefit, highlighting a substantial disparity between the two systems in terms of monthly payouts.

This comparison provides valuable insight into the financial implications of selecting between the OPR and TRS systems, with the TRS system offering higher monthly benefits. This can serve as a crucial factor for decision-making in retirement planning.



## Fixed-Rate Analysis

To modify the simulation and project an employee's pension benefit (TRS) or withdrawal amount (ORP) from retirement until death, we need to consider several key factors:

1. **Retirement Age**: The age at which the employee retires and begins receiving benefits.
2. **Death Age**: The estimated age at which the employee will pass away.


```{r}
# Parameters
retirement_age <- 65
death_age <- 90  # Estimated death age
years_in_retirement <- death_age - retirement_age

# Initialize the mean weighted performance
mean_performance <- mean(
  (equity_us_alloc * yearly_performance_data$US_Equity_Performance) +
  (equity_international_alloc * yearly_performance_data$International_Equity_Performance) +
  (bonds_alloc * yearly_performance_data$Bonds_Performance),
  na.rm = TRUE
)

# Use this value as your fixed return rate for ORP
fixed_return_rate <- (mean_performance-1) #based on the performance of from the last 15 years of investments. 

# Fixed rate assumptions
fixed_return_rate <- fixed_return_rate  # Fixed annual return rate for ORP
withdrawal_rate <- 0.04  # Initial ORP withdrawal rate

# Calculate the average inflation from the combined_data dataset
avg_inflation <- mean(combined_data$inflation_rate, na.rm = TRUE) / 100  # Convert to decimal form


# Starting values
trs_annual_benefit <- annual_benefit_trs  # TRS annual benefit
orp_balance <- yearly_performance_data$Total_Accumulated[years_of_service]  # ORP final balance at retirement

# Initialize vectors to store results
trs_benefits <- numeric(years_in_retirement)
orp_withdrawals <- numeric(years_in_retirement)
orp_balances <- numeric(years_in_retirement)
income_gaps <- numeric(years_in_retirement)

# Simulation
for (year in 1:years_in_retirement) {
  # TRS Benefit (adjusted for inflation with caps)
  if (year == 1) {
    trs_benefits[year] <- trs_annual_benefit
  } else {
    cpi_adjustment <- max(min(avg_inflation * 0.5, 0.03), 0.01)  # Cap between 1% and 3%
    trs_benefits[year] <- trs_benefits[year - 1] * (1 + cpi_adjustment)
  }
  
  # ORP Withdrawal: Fixed percentage of the previous year's balance
  if (year == 1) {
    orp_withdrawals[year] <- orp_balance * withdrawal_rate
    orp_balance <- orp_balance * (1 + fixed_return_rate) - orp_withdrawals[year]
  } else {
    orp_withdrawals[year] <- orp_balances[year - 1] * withdrawal_rate
    orp_balance <- (orp_balances[year - 1] - orp_withdrawals[year]) * (1 + fixed_return_rate)
  }
  
  # Ensure balance never goes negative
  orp_balances[year] <- max(orp_balance, 0)
  if (orp_balance <= 0) {
    orp_withdrawals[year] <- 0
  }
  
  # Income gap (Annually)
  income_gaps[year] <- trs_benefits[year] - orp_withdrawals[year]
}

# Create results data frame
results <- data.frame(
  Year = retirement_age:(death_age - 1),
  TRS_Annual_Income = trs_benefits,
  ORP_Annual_Withdrawal = orp_withdrawals,
  ORP_Balance_Remaining = orp_balances,
  Income_Gap_Annual = income_gaps
)

# Prepare data for visualization
graph_data <- results |>
  select(Year, TRS_Annual_Income, ORP_Annual_Withdrawal, ORP_Balance_Remaining) |>
  pivot_longer(
    cols = c(TRS_Annual_Income, ORP_Annual_Withdrawal),
    names_to = "Plan",
    values_to = "Annual_Income"
  ) |>
  mutate(
    Plan = recode(
      Plan, 
      TRS_Annual_Income = "TRS Annual Income", 
      ORP_Annual_Withdrawal = "ORP Annual Withdrawal"
    )
  )

# Plot TRS and ORP incomes over time
income_plot <- graph_data |>
  ggplot(aes(x = Year, y = Annual_Income, color = Plan, group = Plan)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Comparison of TRS and ORP Annual Income Over Time",
    x = "Year of Retirement",
    y = "Annual Income ($)",
    color = "Retirement Plan"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 14)
  ) +
  scale_y_continuous(labels = scales::dollar_format())

# Display the plots
income_plot
```
## Comparison of TRS and ORP Income

1. **TRS Income**:
   - TRS income grows steadily, driven by inflation adjustments, ensuring higher and more consistent income throughout retirement.

2. **ORP Withdrawals**:
   - ORP withdrawals grow modestly and lack inflation adjustments, leading to slower income growth over time.

3. **Income Gap**:
   - The gap between TRS and ORP widens significantly, with TRS consistently providing higher income.

4. **Key Insight**:
   - TRS is better for retirees needing stable, inflation-protected income, while ORP offers lower growth and purchasing power over time.

```{r}
# Plot ORP remaining balance over time
orp_balance_plot <- results |>
  ggplot(aes(x = Year, y = ORP_Balance_Remaining)) +
  geom_line(color = "blue", size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Growth of ORP Investment Balance Over Time",
    x = "Year of Retirement",
    y = "ORP Balance Remaining ($)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 14)
  ) +
  scale_y_continuous(labels = scales::dollar_format())

# Display the ORP balance plot
orp_balance_plot
```
While TRS may provide a higher and more stable income, the ORP plan demonstrates strong growth in its investment balance over time. This highlights the plan’s long-term sustainability. Additionally, ORP offers greater flexibility, giving employees more control over their investments. 

One key advantage of ORP is that employees can access the full balance if needed, providing the freedom to use the funds as desired, whether for unexpected expenses, large purchases, or other financial goals. This flexibility makes ORP a viable option for those who value control over their retirement savings.

# Monte Carlo Simulation for ORP Plan Retirement Benefits

Monte Carlo simulation is a powerful tool to model the range of possible outcomes for the ORP (Optional Retirement Plan) benefits. By generating multiple scenarios based on random samples of historical data (created using bootstrapping), this approach allows us to evaluate the robustness and performance of the ORP plan under varying conditions.

## **Simulation Overview**
We aim to assess the impact of market conditions and withdrawal rates on the ORP balance over time and evaluate the plan's sustainability under different scenarios.

### **Key Considerations**
- **Death Age**: 90 years  
- **Retirement Age**: 65 years  
- **Withdrawal Rates**: 3% to 10%  
  - We will evaluate the impact of different withdrawal rates on the ORP balance over time.  
- **Number of Simulations**: 200  
  - This ensures a comprehensive analysis by capturing a wide range of potential outcomes.

### **Objectives**
1. **Simulate Market Conditions**:  
   Use bootstrapped historical data to reflect various possible market scenarios.
2. **Evaluate Withdrawal Strategies**:  
   Analyze how different withdrawal rates impact the sustainability of the ORP plan.
3. **Assess Performance**:  
   Provide insights into the range of possible retirement benefits and identify risks or trends under varying scenarios.

This analysis will allow for a better understanding of the ORP plan's robustness and help in making informed decisions for financial planning in retirement.


```{r}

# Parameters
retirement_age <- 65  # Retirement start age
death_age <- 90       # End of simulation age
years_in_retirement <- death_age - retirement_age

# Step 1: Initial ORP Balance
orp_balance <- yearly_performance_data$Total_Accumulated[nrow(yearly_performance_data)]
if (orp_balance == 0 || is.na(orp_balance)) stop("ERROR: ORP Balance is zero or NA.")
# Portfolio allocations
equity_us_alloc <- 0.34
equity_international_alloc <- 0.23
bonds_alloc <- 0.43
if ((equity_us_alloc + equity_international_alloc + bonds_alloc) != 1) stop("ERROR: Allocations must sum to 1.")

# Step 2: Bootstrap CPI Growth
september_cpi <- combined_data$inflation_rate[month(combined_data$date) == 9]
cpi_growth <- september_cpi[-1] / september_cpi[-length(september_cpi)] - 1
set.seed(42)
sampled_inflation <- replicate(200, sample(cpi_growth, years_in_retirement, replace = TRUE))

# Step 3: Sample Returns
sampled_us_returns <- replicate(200, sample(yearly_performance_data$US_Equity_Performance, years_in_retirement, replace = TRUE))
sampled_intl_returns <- replicate(200, sample(yearly_performance_data$International_Equity_Performance, years_in_retirement, replace = TRUE))
sampled_bond_returns <- replicate(200, sample(yearly_performance_data$Bonds_Performance, years_in_retirement, replace = TRUE))

# Step 4: Simulate for Withdrawal Rates (3% to 10%)
withdrawal_rates <- seq(0.03, 0.10, by = 0.01)


trs_benefits <- matrix(0, nrow = 200, ncol = years_in_retirement)

# Initialize storage for all results
all_withdrawal_results <- list()
all_balance_results <- list()  # To store ORP balances for each withdrawal rate

for (rate in withdrawal_rates) {
  orp_withdrawals <- matrix(0, nrow = 200, ncol = years_in_retirement)  # Store withdrawals
  orp_balances <- matrix(0, nrow = 200, ncol = years_in_retirement)    # Store balances for each withdrawal rate
  
  for (sim in 1:200) {
    orp_sim_balance <- orp_balance  # Reset balance for each simulation
    
    for (year in 1:years_in_retirement) {
      # TRS Benefit Adjustment
      if (year == 1) {
        trs_benefit <- trs_annual_benefit
      } else {
        inflation_adjustment <- min(max(sampled_inflation[year, sim] * 0.5, 0.01), 0.03)
        trs_benefit <- trs_benefit * (1 + inflation_adjustment)
      }
      trs_benefits[sim, year] <- trs_benefit

      # ORP Withdrawal Calculation
      withdrawal <- orp_sim_balance * rate
      orp_withdrawals[sim, year] <- withdrawal
      remaining_balance <- max(orp_sim_balance - withdrawal, 0)
      orp_balances[sim, year] <- remaining_balance  # Save balance for the simulation
      
      if (remaining_balance <= 0) break  # Stop simulation if balance is depleted

      # Apply Weighted Returns
      weighted_return <- (equity_us_alloc * (sampled_us_returns[year, sim] - 1)) +
                         (equity_international_alloc * (sampled_intl_returns[year, sim] - 1)) +
                         (bonds_alloc * (sampled_bond_returns[year, sim] - 1))
      orp_sim_balance <- remaining_balance * (1 + weighted_return)
    }
  }

  # Store Average Withdrawals
  avg_withdrawals <- colMeans(orp_withdrawals, na.rm = TRUE)
  result <- data.frame(
    Year = retirement_age:(death_age - 1),
    Average_Withdrawal = avg_withdrawals,
    Withdrawal_Rate = paste0(round(rate * 100, 1), "%")
  )
  all_withdrawal_results[[paste0("Rate_", round(rate * 100, 1))]] <- result
  
  # Store the ORP Balances for the Current Rate
  all_balance_results[[paste0("Rate_", round(rate * 100, 1))]] <- orp_balances
}

# Combine Withdrawal Results for Visualization
combined_withdrawal_results <- do.call(rbind, all_withdrawal_results)

# Add TRS Results
combined_trs_results <- data.frame(
  Year = retirement_age:(death_age - 1),
  Average_Withdrawal = colMeans(trs_benefits, na.rm = TRUE),
  Withdrawal_Rate = "TRS Benefit",
  Line_Type = "dashed",
  Legend = "TRS Benefit"
)

# Adjust ORP Results for Consistency
combined_withdrawal_results$Line_Type <- "solid"
combined_withdrawal_results$Legend <- combined_withdrawal_results$Withdrawal_Rate

# Merge Results
final_combined_results <- rbind(combined_withdrawal_results, combined_trs_results)
final_combined_results$Legend <- factor(
  final_combined_results$Legend, 
  levels = c("3%", "4%", "5%", "6%", "7%", "8%", "9%", "10%", "TRS Benefit")
)

# Plot Withdrawals
withdrawal_plot <- ggplot(final_combined_results, aes(x = Year, y = Average_Withdrawal, color = Legend, linetype = Line_Type, group = Legend)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("blue", "green", "red", "purple", "orange", "cyan", "brown", "pink", "black")) +
  scale_linetype_identity() +
  labs(
    title = "Comparison of ORP Withdrawals and TRS Benefits",
    x = "Year of Retirement",
    y = "Average Withdrawal/Benefit ($)",
    color = "Legend"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = scales::dollar_format())

# Combine Balance Results for Visualization
combined_balance_results <- do.call(rbind, lapply(seq_along(all_balance_results), function(i) {
  avg_balance <- colMeans(all_balance_results[[i]], na.rm = TRUE)  # Compute average balances
  data.frame(
    Year = retirement_age:(death_age - 1),
    Average_Balance = avg_balance,
    Withdrawal_Rate = paste0(round(withdrawal_rates[i] * 100, 1), "%")  # Match withdrawal rate
  )
}))

# Plot Balances
balance_plot <- ggplot(combined_balance_results, aes(x = Year, y = Average_Balance, color = Withdrawal_Rate, group = Withdrawal_Rate)) +
  geom_line(size = 1.2) +
  scale_color_manual(values = c("blue", "green", "red", "purple", "orange", "cyan", "brown", "pink")) +
  labs(
    title = "Comparison of ORP Balances Across Withdrawal Rates",
    x = "Year of Retirement",
    y = "Average ORP Balance ($)",
    color = "Withdrawal Rate"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = scales::dollar_format())


# Display Plots
withdrawal_plot
```


We clearly see that the TRS is a better model for retirement. The TRS benefit remains stable and increases over time due to inflation adjustments, providing a reliable income source for retirees. Also, we decided to compare the ORP withdrawals at different rates. Appreciating a huge gap just by changing one percent. Making it a clear factor when comes down to retirement planning. 

```{r}
balance_plot
```

Also, we want to see the average balance of the ORP plan over time. The line graph illustrates the average ORP balances across different withdrawal rates over the retirement period. Key insights from the visualization are as follows:

1. **Withdrawal Rates**:
   - The graph shows the average ORP balance for withdrawal rates ranging from 3% to 10%.
   - The 6% withdrawal rate maintains a stable balance over time, providing a consistent income stream for retirees, and still increase its value.
   - The 10% withdrawal rate depletes the ORP savings too quickly, indicating a high-risk strategy.
   - The 3% withdrawal rate is overly conservative, resulting in a slow depletion of savings and lower income.
   
This plan is recommended for those who want to have a stable income over time and may want to liquidate in the future or pass it on to their partners or children. 


After getting this information, we want to ensure that the ORP plan is sustainable and will not run out of money before the retiree passes away. This ensures that the retiree will have a stable income over time.
 

```{r}

# Split orp_balances into a list based on withdrawal rates
orp_balances <- split(orp_balances, rep(1:length(withdrawal_rates), each = years_in_retirement))

# Convert each list element to a matrix for simulation
orp_balances <- lapply(orp_balances, function(x) matrix(x, nrow = 200, ncol = years_in_retirement))

# Calculate probabilities of exhaustion for each withdrawal rate
exhaust_probabilities <- sapply(seq_along(withdrawal_rates), function(rate_index) {
  current_balance <- orp_balances[[rate_index]]  # Get balances for the current rate
  exhausted <- apply(current_balance, 1, function(sim_row) any(sim_row[1:(years_in_retirement - 1)] == 0))  # Check if exhausted
  mean(exhausted) * 100  # Return percentage of exhausted simulations
})

# Create a data frame for results
exhaustion_table <- data.frame(
  Withdrawal_Rate = paste0(round(withdrawal_rates * 100, 1), "%"),
  Probability = as.numeric(round(exhaust_probabilities, 2))
)

# Load the gt library for table formatting
library(gt)

# Generate a formatted table with results
gt_exhaustion_table <- exhaustion_table |>
  gt() |>
  tab_header(
    title = "Probability of ORP Savings Exhaustion",
    subtitle = "Before Death by Withdrawal Rate"
  ) |>
  fmt_number(
    columns = c(Probability),  # Updated syntax
    decimals = 2
  ) |>
  cols_label(
    Withdrawal_Rate = "Withdrawal Rate",
    Probability = "Exhaustion Probability (%)"
  ) |>
  tab_source_note(
    source_note = "Data simulated using Monte Carlo bootstrapping."
  )

# Display the table
gt_exhaustion_table
```
In all cases, the ORP plan is sustainable. However, we have to take into consideration that the models with a that 7 percent, would have a negative growth in terms of withdrawals. This means that the retiree would have to adjust the withdrawal rate to a lower percentage to make sure that the ORP plan is sustainable. 

We want to keep digging on the ORP plan and see if the retiree would have a higher income than the TRS plan. 

```{r}
# Initialize vector to store probabilities for each withdrawal rate
higher_income_probabilities <- numeric(length(withdrawal_rates))

# Loop through withdrawal rates
for (i in seq_along(withdrawal_rates)) {
  rate <- withdrawal_rates[i]
  higher_income_count <- 0
  
  # Iterate over years
  for (year in 1:years_in_retirement) {
    higher_income_count <- higher_income_count + sum(orp_withdrawals[, year] > trs_benefits[, year])
  }
  
  # Calculate probability
  higher_income_probabilities[i] <- higher_income_count / (200 * years_in_retirement) * 100
}

# Create a table
higher_income_table <- data.frame(
  Withdrawal_Rate = paste0(round(withdrawal_rates * 100, 1), "%"),
  Probability = round(higher_income_probabilities, 2)
)

# Display the table
gt <- gt(higher_income_table) |>
  tab_header(
    title = "Probability of Higher ORP Income vs. TRS Benefit",
    subtitle = "Based on Different Withdrawal Rates"
  ) |>
  cols_label(
    Withdrawal_Rate = "Withdrawal Rate",
    Probability = "Higher Income Probability (%)"
  ) |>
  tab_spanner(label = "Higher Income Probability (%)", columns = c(2)) |>
  tab_options(
    table.font.size = "medium",
    heading.align = "center"
  )
gt
```
As we can see, the percentage doesn't change on all the withdrawal rates. This means that the OPR could have better months than the TRS. However, as we saw in the previous graphs, that just would happens if the retiree has a higher withdrawal rate and would just happens for the first few years. 

# Final Recommendations for OPR

Now we want to prove why our recommendation is 6% for the withdrawal rate. 

```{r}
library(gt)

# Filter data for the 6% withdrawal rate
rate_6_data <- combined_withdrawal_results[combined_withdrawal_results$Withdrawal_Rate == "6%", ]
rate_6_exhaust_prob <- exhaustion_table[which(withdrawal_rates == 0.06), "Probability"]

# Add TRS Results as a Data Frame
trs_data <- data.frame(
  Year = retirement_age:(death_age - 1),
  Average_Withdrawal = colMeans(trs_benefits, na.rm = TRUE),
  Legend = "TRS Benefit"
)

# Calculate Initial and Final Income for Each Program
# For ORP (6% Withdrawal Rate)
orp_initial_income <- rate_6_data$Average_Withdrawal[1]  # First year withdrawal
orp_income_death <- tail(rate_6_data$Average_Withdrawal, 1)  # Last year withdrawal
orp_difference <- orp_income_death - orp_initial_income

# For TRS
trs_initial_income <- trs_data$Average_Withdrawal[1]  # First year benefit
trs_income_death <- tail(trs_data$Average_Withdrawal, 1)  # Last year benefit
trs_difference <- trs_income_death - trs_initial_income

# Create the Data Frame for the Table
income_comparison_table <- data.frame(
  Plan = c("ORP (6% Withdrawal Rate)", "TRS Benefit"),
  Initial_Income = c(round(orp_initial_income, 2), round(trs_initial_income, 2)),
  Income_at_Death = c(round(orp_income_death, 2), round(trs_income_death, 2))
)

# Generate the Table Using `gt`
income_comparison_table |>
  gt() |>
  tab_header(
    title = "Income Comparison Between ORP and TRS",
    subtitle = "Initial and Final Income with Differences at Death"
  ) |>
  cols_label(
    Plan = "Plan",
    Initial_Income = "Initial Income ($)",
    Income_at_Death = "Income at Death ($)",
  ) |>
  fmt_currency(
    columns = c(Initial_Income, Income_at_Death),
    currency = "USD"
  ) |>
  tab_options(
    table.font.size = "medium",
    heading.title.font.size = "large",
    heading.subtitle.font.size = "medium"
  )


# Visualization: Compare 6% ORP withdrawals with TRS benefits
ggplot() +
  geom_line(data = rate_6_data, aes(x = Year, y = Average_Withdrawal, color = "6% Withdrawal Rate"), size = 1.2) +
  geom_line(data = trs_data, aes(x = Year, y = Average_Withdrawal, color = Legend), linetype = "dashed", size = 1.2) +
  scale_color_manual(
    values = c("6% Withdrawal Rate" = "blue", "TRS Benefit" = "black"),
    labels = c("6% Withdrawal Rate", "TRS Benefit")
  ) +
  labs(
    title = "Comparison of 6% ORP Withdrawals and TRS Benefits",
    x = "Year of Retirement",
    y = "Average Withdrawal/Benefit ($)",
    color = "Legend"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 14)
  ) +
  scale_y_continuous(labels = scales::dollar_format())

```

After analyzing the ORP plan and comparing it with the TRS benefits, we can provide the following recommendations based on the simulation results:
- The annual income of the retiree would be highly affected by the withdrawal rate. A 6% withdrawal rate would provide a stable income over time, ensuring the sustainability of the ORP plan. It would kind of simulate the TRS benefits in terms of growth and stability. But with a certain difference in income, to start, TRS has ~ 5 thousand of difference, and at the end, it has ~ 7  thousand of difference. But still is a the best option for the retiree.


# Conclusions

The analysis highlights that the Teachers Retirement System (TRS) offers superior income stability and inflation protection compared to the Optional Retirement Plan (ORP). TRS consistently provides higher annual income, starting at ~$1,769/month versus ORP’s ~$883/month (with the common suggested 4%), with the gap widening over time to ~$7,000 annually by the end of retirement.

The ORP plan, while offering flexibility and growth through compounding, requires careful management of withdrawal rates. A 6% withdrawal rate strikes the best balance, ensuring sustainability and consistent income. However, ORP withdrawals are not inflation-adjusted, leading to reduced purchasing power over time.

Recommendation:
Choose TRS for stable, inflation-protected income.
Opt for ORP if you value investment control and can manage withdrawal rates responsibly and if you want to leave money for your family.
TRS remains the optimal choice for most retirees seeking long-term financial security.