---
title: "STA/OPR 9750 Mini-Project #02: The Business of Show Business"
author: "Juan Camilo Martinez Benavides"
date: today
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    theme: darkly
    embed-resources: true
execute:
  warning: false
  message: false
---

## Executive Summary

This project analyzes factors influencing movie success using IMDb data, focusing on ratings, audience engagement, and revenue. We identify key trends in genre performance over decades and develop a predictive model for film success. The findings culminate in a data-backed proposal for a remake of *North by Northwest*.

## Introduction

This project defines what makes a movie successful by analyzing IMDb ratings, audience engagement, and revenue to create a predictive model. We uncover trends in genre evolution, offering insights for future productions.

## Data Source

For this project, we utilize data from the [Internet Movie Database (IMDb)](https://datasets.imdbws.com), a widely recognized source of comprehensive movie information. This data is freely available for non-commercial use, offering a rich foundation for analyzing film success factors and industry trends.

The following code downloads the IMDb datasets into ***R*** for analysis:

```{r}
#| echo: true
#| message: false
#| warning: false
get_imdb_file <- function(fname){
    BASE_URL <- "https://datasets.imdbws.com/"
    fname_ext <- paste0(fname, ".tsv.gz")
    if(!file.exists(fname_ext)){
        FILE_URL <- paste0(BASE_URL, fname_ext)
        download.file(FILE_URL, 
                      destfile = fname_ext)
    }
    as.data.frame(readr::read_tsv(fname_ext, lazy=FALSE))
}

NAME_BASICS <- get_imdb_file("name.basics")
```

For readability, we change the name of the following tables:

```{r}
#| echo: true
#| message: false
#| warning: false
TITLE_BASICS     <- get_imdb_file("title.basics")
TITLE_EPISODES   <- get_imdb_file("title.episode")
TITLE_RATINGS    <- get_imdb_file("title.ratings")
TITLE_CREW       <- get_imdb_file("title.crew")
TITLE_PRINCIPALS <- get_imdb_file("title.principals")
```

Due to the large size of these files, loading them may take some time. To speed up the process, you can cache the code chunk that reads the files, which will prevent reloading them every time you run the code.

## Data Sub-Sampling

Since the data set contains a vast number of data points, we will narrow it down to ensure smooth analysis. For the ***NAME_BASICS*** table, we will focus on individuals with at least two "known for" credits, allowing us to work with a more manageable subset of relevant data.

```{r}
#| echo: true
#| message: false
#| warning: false
library(dplyr)
library(stringr)

NAME_BASICS <- NAME_BASICS |> 
    filter(str_count(knownForTitles, ",") > 1)

```

IMDb has a long tail of obscure movies:

```{r}
#| echo: true
#| message: false
#| warning: false
library(ggplot2)
library(scales)

TITLE_RATINGS |>
    ggplot(aes(x=numVotes)) + 
    geom_histogram(bins=30) +
    xlab("Number of IMDB Ratings") +
    ylab("Number of Titles") + 
    ggtitle("Majority of IMDB Titles Have Less than 100 Ratings") + 
    theme_bw() + 
    scale_x_log10(label=scales::comma) + 
    scale_y_continuous(label=scales::comma)

```

As we can see, the majority of titles have fewer than 100 ratings. Therefore, it would be beneficial to filter out titles with less than 100 ratings to focus on more widely rated films. This becomes even clearer in the following visualization:

```{r}
TITLE_RATINGS |>
    pull(numVotes) |>
    quantile()
```

By applying this drop, we significantly reduce the size of our data set:

```{r}
#| echo: true
#| message: false
#| warning: false
TITLE_RATINGS <- TITLE_RATINGS |>
    filter(numVotes >= 100)

```

Next, we will join the tables to filter out titles with fewer than 100 ratings. We'll use the `semi_join` function to retain only the rows that exist in both tables, ensuring we focus on titles with a higher number of ratings.

```{r}
#| echo: true
#| message: false
#| warning: false
TITLE_BASICS <- TITLE_BASICS |>
    semi_join(TITLE_RATINGS, 
              join_by(tconst == tconst))

TITLE_CREW <- TITLE_CREW |>
    semi_join(TITLE_RATINGS, 
              join_by(tconst == tconst))

TITLE_EPISODES_1 <- TITLE_EPISODES |>
    semi_join(TITLE_RATINGS, 
              join_by(tconst == tconst))
TITLE_EPISODES_2 <- TITLE_EPISODES |>
    semi_join(TITLE_RATINGS, 
              join_by(parentTconst == tconst))

TITLE_EPISODES <- bind_rows(TITLE_EPISODES_1,
                            TITLE_EPISODES_2) |>
    distinct()

TITLE_PRINCIPALS <- TITLE_PRINCIPALS |>
    semi_join(TITLE_RATINGS, join_by(tconst == tconst))


rm(TITLE_EPISODES_1)
rm(TITLE_EPISODES_2)

```

## Taking a Look at the Data

We will first inspect the data to determine if any changes to the data types are necessary. In this dataset, most columns are being read as character types. This typically happens when missing values are represented in a non-standard way. For example, in these files, missing values are represented as ***\\N***. Since ***R*** does not automatically recognize this as an ***NA*** value, it treats them as strings.

To address this, we need to:

1.  Use the `mutate` function to modify the columns.
2.  Apply the `as.numeric` function to convert columns to the correct data type.

We can clean the datasets using the following code:

```{r}
#| echo: true
#| message: false
#| warning: false
NAME_BASICS <- NAME_BASICS |>
    mutate_if(is.character, ~na_if(., "\\N")) |>
    mutate(birthYear = as.numeric(birthYear),
           deathYear = as.numeric(deathYear))
```

```{r}

TITLE_BASICS <- TITLE_BASICS |>
    mutate_if(is.character, ~na_if(., "\\N")) |>
    mutate(
        startYear = as.numeric(startYear), 
        endYear = as.numeric(endYear),    
        runtimeMinutes = as.numeric(runtimeMinutes)  
    )

```

```{r}
TITLE_EPISODES <- TITLE_EPISODES |>
    mutate_if(is.character, ~na_if(., "\\N")) |>
    mutate(
        seasonNumber = as.numeric(seasonNumber),
        episodeNumber = as.numeric(episodeNumber)
    )

```

```{r}
TITLE_RATINGS <- TITLE_RATINGS |>
    mutate_if(is.character, ~na_if(., "\\N")) |>
    mutate(
        averageRating = as.numeric(averageRating), 
        numVotes = as.numeric(numVotes)  
    )
```

```{r}
TITLE_CREW <- TITLE_CREW |>
    mutate_if(is.character, ~na_if(., "\\N"))
```

```{r}
TITLE_PRINCIPALS <- TITLE_PRINCIPALS |>
    mutate_if(is.character, ~na_if(., "\\N")) |>
    mutate(characters = str_replace_all(characters, '\\[|\\"|\\]', '')) #clean undesired characters

```

During the cleaning process, we transformed string columns into numeric columns and cleaned the character columns in the ***TITLE_PRINCIPALS*** table. Additionally, across all tables, we replaced any occurrences of ***\\N*** with ***NA*** values, allowing us to handle missing data more efficiently. In the ***TITLE_PRINCIPALS*** table, we also removed unwanted characters to enhance data quality.

By cleaning and preparing the data, we now have a clearer view of the relationships between tables and the variables at our disposal.

## Data Analysis

With the data now properly cleaned and filtered, we can move forward with our analysis. To begin, we will address the following key questions:

1.  How many movies are in our dataset?
2.  How many TV series?
3.  How many TV episodes?

Answering these questions will help us gain an overview of the data set and set the foundation for deeper insights.

```{r}
#| echo: true
#| message: false
#| warning: false
# Install and load the gt package
if(!require(gt)) {
  install.packages("gt")
}
library(gt)

# Get the count of movies, TV series, and TV episodes
movie <- TITLE_BASICS |>
  filter(titleType == "movie") |>
  nrow()

tvSeries <- TITLE_BASICS |>
  filter(titleType == "tvSeries") |>
  nrow()

tvEpisode <- TITLE_BASICS |>
  filter(titleType == "tvEpisode") |>
  nrow()

# Create a summary table and apply gt
summary_table <- tibble(
  Title_Type = c("Movies", "TV Series", "TV Episode"),
  Count = c(movie, tvSeries, tvEpisode)
) |>
  gt() |>
  tab_header(
    title = "Number of Movies, TV Series, and TV Episodes"
  )
summary_table

```

We observed a notable trend in our dataset: the number of movies and TV episodes is quite similar, while TV series have the lowest count among the three categories. This insight is particularly interesting as it may guide future analyses and decision-making in the film industry. Understanding these dynamics can help inform production strategies, audience engagement initiatives, and content development.

## Data Exploration: Oldest Living Person

To identify the oldest living person in our dataset, we can focus on the ***NAME_BASICS*** table, filtering by individuals who are still alive and sorting by birth year. This will allow us to determine who holds the title of the oldest living person in the film industry.

```{r}
library(dplyr)
library(DT)  # Use DT for pagination

oldest_person <- NAME_BASICS |>         
  filter(is.na(deathYear)) |>            # Keep only people with NA in deathYear (still alive)
  mutate(age = 2024 - birthYear) |>      # Calculate their age in 2024
  filter(age > 110) |> 
  filter(age < 116)  # Filter for ages below 117 year  # Find the person with the maximum age

summary_table <- oldest_person |> 
  select(Name = primaryName, "Year of Birth" = birthYear, Age = age)

datatable(summary_table, 
          options = list(pageLength = 5),   # Limit to 5 rows per page
          caption = 'Oldest Person Alive in this Data Set')


```

Initially, we compiled a list of individuals who were 117 years old based on the dataset. However, after further investigation, we found that none of them were alive. We then narrowed the search to people aged 100 to 116, where we identified 297 individuals still living, aided by external sources such as the Internet. Despite this, finding definitive information about the oldest living person remains a challenge, as much of this data is not readily available.

## Identifying a Perfectly Rated TV Episode

Next, we aimed to identify a TV episode with a perfect 10/10 IMDb rating and at least 200,000 ratings. By filtering our dataset, we can uncover this exceptional episode and determine the series it belongs to.

```{r}

good_ranking <- TITLE_BASICS |>
  inner_join(TITLE_RATINGS, by = "tconst") |>  # Join to bring in titleType from TITLE_BASICS
  filter(titleType == "tvEpisode") |>        # Now you can filter for TV episodes
  filter(averageRating == 10 & numVotes > 200000) |>  # Filter for 10/10 rating and >200k votes
  arrange(desc(numVotes))  # Sort in descending order by number of votes

# Create the summary table
summary_table <- good_ranking |>
  select(Name = primaryTitle, rating = averageRating, "Number of Votes" = numVotes ) |> 
  gt() |>                       
  tab_header(
    title = "TV Episode with a Rating of 10/10 and More Than 200,000 Votes"
  )

summary_table
```

Our analysis revealed that the **Breaking Bad** episode *"Ozymandias"* has achieved a perfect **10/10 rating** with over **229,000 votes** on IMDb. This episode stands as a milestone in television history, showcasing the pinnacle of storytelling and character development.

## Exploring Mark Hamill's Most Known Projects

We will now explore the four projects that actor **Mark Hamill** is most known for on IMDb. Best recognized for his role as **Luke Skywalker** and his voice work as **The Joker**, Hamill's career spans decades. This exploration will help identify the roles that have shaped his legacy across both film and animation.

```{r}

mark_hamill <- NAME_BASICS |>
  filter(primaryName == "Mark Hamill") |>
  select(knownForTitles)

tconsts <- strsplit(mark_hamill$knownForTitles, ",")[[1]]

mark_hamill_title <- TITLE_BASICS |>
   filter(tconst %in% tconsts) |>
   select("Title Name" = primaryTitle, Type = titleType) |>
   gt() |>
    tab_header(
    title = "The four Projects Mark Hamill Is Best Known For")


mark_hamill_title
```

From our analysis, we can see that Mark Hamill's top four "Most Known For" projects are all from the *Star Wars* franchise, solidifying his iconic role as **Luke Skywalker**. These films have defined his career and continue to resonate with audiences across generations.

## Next Exploration: Highest Rated TV Series

Building on our analysis, we now turn our attention to identifying the **TV series with more than 12 episodes** that has the highest average rating. By exploring this, we aim to discover which series has consistently captivated audiences over a longer run.

```{r}
series <- TITLE_BASICS |>
  inner_join(TITLE_EPISODES, by = "tconst") |>
  inner_join(TITLE_RATINGS,by = "tconst")

best_series <- series |>
  filter(episodeNumber > 12) |>
  distinct() |>
  filter(averageRating == max(averageRating, na.rm = TRUE)) |>
  select("Tile Name" = primaryTitle, "Average Ratings" = averageRating) |>
  gt() |>
  tab_header(
    title = "The TV Series with the Highest Average Rating and More Than 12 Episodes"
  )
  
best_series

```

As seen in the results, several series with more than 12 episodes hold a perfect rating of 10. However, duplicates such as "Bolum" (likely individual episodes incorrectly listed) indicate inconsistencies in the source data.

## Investigating "Jump the Shark" in *Happy Days*

The idiom "jump the shark" originates from a 1977 *Happy Days* episode where a character jumps over a shark on water skis, marking a perceived decline in quality. We investigate whether this cultural moment actually corresponds to a drop in viewer ratings.

To investigate this, we now ask: Is it true that episodes from the later seasons of *Happy Days* have lower average ratings than the early seasons? Let's explore the data to see if there's a noticeable drop in ratings as the series progressed.

```{r}

data <- TITLE_BASICS |>
  inner_join(TITLE_EPISODES, join_by(tconst == parentTconst)) |>  # Join episodes with series
  inner_join(TITLE_RATINGS, by = "tconst")  # Join ratings dat

happy_days <- data |>
    filter(primaryTitle == "Happy Days") |>
    arrange(seasonNumber) |>
    select(primaryTitle, "Season Number" = seasonNumber, "Average Rating" = averageRating, year = endYear) |>
    distinct() |>

  gt() |>
  tab_header(
    title = "Happy Days Seasons & Ratings per season"
  )

happy_days
```

The assumption that the later seasons of ***Happy Days*** had lower average ratings is not true. As we can observe, the average rating remained constant at **7.4** across all seasons. This disproves the hypothesis that the 1977 "jump the shark" event had a significant impact on the show's overall ratings.

Despite the controversy surrounding that episode, the data shows no evidence of a decline in audience ratings afterward.

## Creating Criteria for How Successful a Movie Is:

As we aim to propose a new movie, it is important to understand how successful it might be. While success is often subjective, we are using data from the **TITLE_RATING** table to measure success, considering **average rating** as a measure of quality and **number of votes** as a measure of popularity.

Traditionally, a movie's success is also evaluated by the revenue it generates. Therefore, we have selected **seven movies with the highest revenue** and **five movies that lost money**, with data sourced from [Box Office Mojo](https://www.boxofficemojo.com).

## Movie Revenue Table

| Successful Movies and Revenue | Unsuccessful Movies, Money Lost |
|--------------------------------------|----------------------------------|
| Avatar (\$2.92 billion) | John Carter (-\$0.20 billion) |
| Avatar: The Way of Water (\$2.32 billion) | The Lone Ranger (-\$0.19 billion) |
| Titanic (\$2.25 billion) | Mars Needs Moms (-\$0.15 billion) |
| Star Wars: Episode VII - The Force Awakens (\$2.07 billion) | King Arthur: Legend of the Sword (-\$0.15 billion) |
| Spider-Man: No Way Home (\$1.92 billion) | Sinbad: Legend of the Seven Seas (-\$0.125 billion) |
| Jurassic World (\$1.67 billion) | Cutthroat Island (-\$0.1 billion) |
| Top Gun: Maverick (\$1.49 billion) | R.I.P.D. (-\$0.1 billion) |

```{r}
#first we are filtering the movies we need, selecting the usuful tables, such as originalTitle, averageRating, numVotes
movies <- TITLE_BASICS |>
  inner_join(TITLE_RATINGS, by = "tconst") |>
  filter(titleType == "movie")
  
filter_movies <- movies |>
  filter((str_detect(str_to_lower(primaryTitle), "avatar") & startYear == 2009) |
          (str_detect(str_to_lower(primaryTitle), "avatar: the way of water") & startYear == 2022) |
          (primaryTitle == "Titanic" & startYear == 1997) |
          (str_detect(str_to_lower(primaryTitle), "star wars") & startYear == 2015) |
          (str_detect(str_to_lower(primaryTitle), "spider-man: no way home") & startYear == 2021) |
          (str_detect(str_to_lower(primaryTitle), "jurassic world") & startYear == 2015) |
          (str_detect(str_to_lower(primaryTitle), "top gun: maverick") & startYear == 2022) |
          (str_detect(str_to_lower(primaryTitle), "john carter")) |
          (str_detect(str_to_lower(primaryTitle), "the lone ranger") & startYear == 2013) |
          (str_detect(str_to_lower(primaryTitle), "mars needs moms")) |
          (str_detect(str_to_lower(primaryTitle), "king arthur") & startYear == 2017) |
          (primaryTitle ==  "Cutthroat Island") |
          (primaryTitle ==  "R.I.P.D.") |
          (str_detect(str_to_lower(primaryTitle), "legend of the seven seas")))
filter_movies <- filter_movies |>
  arrange(primaryTitle)

# Assign revenue to the filtered dataframe
filter_movies$Revenue <- c(
  2.92,   # Avatar (successful)
  2.32,   # Avatar: The Way of Water (successful)
  -0.10,  # Cutthroat Island (loss)
  -0.20,  # John Carter (loss)
  1.67,   # Jurassic World (successful)
  -0.15,  # King Arthur: Legend of the Sword (loss)
  -0.15,  # Mars Needs Moms (loss)
  -0.10,  # R.I.P.D. (loss)
  -0.125, # Sinbad: Legend of the Seven Seas (loss)
  1.92,   # Spider-Man: No Way Home (successful)
  2.07,   # Star Wars: Episode VII - The Force Awakens (successful)
  -0.19,  # The Lone Ranger (loss)
  2.25,   # Titanic (successful)
  1.49    # Top Gun: Maverick (successful)
)

filter_movies |>
  arrange(desc(Revenue)) |>  # Sort by revenue
  select("Original Name" = primaryTitle, "Average Rating" = averageRating, "Votes" = numVotes, "Revenue (Billion)" = Revenue) |>
  gt() |>
  tab_header(
    title = "7 most Sucessful and 7 most Unsuccessful Movies based on revenue (on Billion)"
   )
```

After adding the revenue column, we will create a model to serve as a criterion for success. The model is based on the following assumptions:

1.  **Scale**: Our success scale will range from 0 to 10, with 10 representing an extremely successful movie.
2.  **Revenue Mapping**: For the positive revenue values (successful movies), we will assign ratings between **9.5 and 10**, indicating extreme success. For negative revenue values (unsuccessful movies), we will assign ratings between **4 and 7**, we choose this number after multiple iterations.

This range was chosen because, as observed in our table, even if a movie receives a large number of votes, ratings below 7 tend to correlate with more negative feedback, the opposite of what we expect from a successful movie, regardless of popularity.

```{r}
# Step 1: Define the max and min values for positive and negative revenues
max_positive_revenue <- max(filter_movies$Revenue[filter_movies$Revenue > 0])
min_positive_revenue <- min(filter_movies$Revenue[filter_movies$Revenue > 0])

max_negative_revenue <- max(filter_movies$Revenue[filter_movies$Revenue < 0])  # closest to 0
min_negative_revenue <- min(filter_movies$Revenue[filter_movies$Revenue < 0])  # most negative

# Step 2: Apply different formulas based on whether revenue is positive or negative
filter_movies$Success_Score <- ifelse(
  filter_movies$Revenue > 0, 
  9.5 + 0.5 * (filter_movies$Revenue - min_positive_revenue) / (max_positive_revenue - min_positive_revenue),  # Positive revenue: 9.5 to 10 scale
  4 +  3 * (filter_movies$Revenue - min_negative_revenue) / (max_negative_revenue - min_negative_revenue)  # Negative revenue: 6 to 7 scale
)

filter_movies |>
  arrange(desc(Success_Score)) |> # Sort by revenue
  select("Original Name" = primaryTitle, "Average Rating" = averageRating, "Votes" = numVotes, "Success Score" = Success_Score ) |>
  gt() |>
  tab_header(
    title = "7 most Sucessful and 7 most Unsuccessful Movies based on Success Score (on Billion)"
   )
```

Based on the assigned success scores, we will apply a **linear regression model** to determine the parameters **B0**, **B1**, and **B2**. These parameters will help us create an equation that can be applied to the rest of the dataset, allowing us to predict the success of other movies based on their features.

The linear regression model will use the following:

-   **B0**: The intercept, representing the baseline success score.
-   **B1**: The coefficient for the average rating, indicating the effect of quality on a movie's success.
-   **B2**: The coefficient for the number of ratings, reflecting how popularity impacts success.

```{r}
#| echo: true
#| message: false
#| warning: false

# Fit the linear regression model
model <- lm(Success_Score ~ averageRating + numVotes, data = filter_movies)

summary(model)
```

Based on the regression analysis, the final model to predict the success score is represented by the following equation:

$$
\text{Success Score} = 2.133 + 0.5842 \times (\text{Average Rating}) + 2.730 \times 10^{-6} \times (\text{Votes})
$$

-   The R-squared value is **0.6283**, meaning that about **62.83%** of the variation in the success score can be explained by the average rating and number of votes.
-   The Adjusted R-squared is **0.5607**, slightly lower, accounting for the number of predictors in the model.
-   With a **p-value of 0.0043**, the overall model is statistically significant.

Based on our analysis, we have determined that the threshold for a movie to be considered successful is an average rating of **7.9 or higher**.

# Validating Our Model

To validate our formula, we will apply it to a set of **14 movies** and evaluate whether they meet the success criteria based on their predicted success scores. This will allow us to assess the accuracy and effectiveness of the model in predicting movie success.

```{r}

# Calculate the new success score based on the formula
filter_movies$SuccessNewScore <- 2.133 + filter_movies$averageRating * 0.5842 + filter_movies$numVotes * 2.730 * 10^-6

# Create a new column indicating "Successful" or "Unsuccessful" based on the threshold of 7.9
filter_movies$SuccessStatus <- ifelse(filter_movies$SuccessNewScore > 7.9, "Successful", "Unsuccessful")

# Select the desired columns and display them using the gt table
filter_movies |>
  select("Original Name" = primaryTitle, 
         "Average Rating" = averageRating, 
         "Votes" = numVotes, 
         "Revenue (Billion)" = Revenue, 
         "Test Success Score" = Success_Score,
         "New Success Score" = SuccessNewScore, 
         "Status" = SuccessStatus) |>
  gt() |>
  tab_header(
    title = "7 Most Successful and 7 Most Unsuccessful Movies Based on Revenue (in Billions)"
  )

```

Based on the analysis of all 14 movies, the model proves to be accurate in predicting a movie's success. The movies that were successful in terms of revenue were also classified as successful by the model, while the movies that lost money were correctly identified as unsuccessful. This demonstrates the effectiveness of the model in using average rating and number of votes to predict movie success.

### Identifying Poorly Rated Popular Movies

We will identify **5 movies** that have a **large number of IMDb votes** but score poorly on our success metric. To do this, we will filter for movies with a **rating below 6** (indicating poor quality) and **more than 100,000 votes**, using the original "movies" dataset. This will allow us to confirm if these movies are indeed of low quality based on their performance on IMDb.

```{r}

TITLE_RATINGS$SuccessNewScore <- 2.133 + TITLE_RATINGS$averageRating * 0.5842 + TITLE_RATINGS$numVotes * 2.730 * 10^-6

TITLE_RATINGS$SuccessStatus <- ifelse(TITLE_RATINGS$SuccessNewScore > 7.9, "Successful", "Unsuccessful")

movies <- TITLE_BASICS |>
  inner_join(TITLE_RATINGS, by = "tconst") |>
  filter(titleType == "movie")


five_movies <- movies |>
  filter(averageRating < 6 & numVotes > 100000) |>
  slice_head(n = 5) |>  # Select the first 5 movies
  select("Original Name" = primaryTitle, 
         "Average Rating" = averageRating, 
         "Votes" = numVotes, 
         "Success Score" = SuccessNewScore, 
         "Status" = SuccessStatus) |>
  gt() |>
  tab_header(
    title = "5 Movies with High Number of Votes and Low Success Scores"
  )

five_movies
```

As we can see, our model correctly classifies movies with a low score and a high number of votes as unsuccessful. This further confirms the accuracy of our model in predicting a movie's success based on its average rating and number of votes.

## Analyzing a Prestige Actor: Leonardo DiCaprio

Next, we will examine the projects of a prestigious actor, **Leonardo DiCaprio**, and confirm that many of his films achieve high scores on our success metric. By applying the model to his body of work, we can validate the consistency of his successful film career and analyze the impact of his films based on our criteria.

```{r}

# Filter the data for Leonardo DiCaprio and Christopher Nolan
leonardo_dicaprio <- NAME_BASICS |>
  filter(primaryName == "Leonardo DiCaprio") |>
  select(knownForTitles)

nolan <- NAME_BASICS |>
  filter(primaryName == "Christopher Nolan") |>
  select(knownForTitles)

# Split the knownForTitles column to get the tconsts
tconsts_leonardo <- strsplit(leonardo_dicaprio$knownForTitles, ",")[[1]]
tconsts_nolan <- strsplit(nolan$knownForTitles, ",")[[1]]

# Filter the movies based on the tconsts for Leonardo DiCaprio and Christopher Nolan
leonardo_movies <- movies |>
  filter(tconst %in% tconsts_leonardo) |>
  mutate(Person = "Leonardo DiCaprio")

nolan_movies <- movies |>
  filter(tconst %in% tconsts_nolan) |>
  mutate(Person = "Christopher Nolan")

leonardo_nolan_movies <- bind_rows(leonardo_movies, nolan_movies)

leonardo_nolan_movies |>
  select("Original Name" = primaryTitle, 
         "Average Rating" = averageRating, 
         "Votes" = numVotes, 
         "Status" = SuccessStatus, 
         "Associated With" = Person) |>
  gt() |>
  tab_header(
    title = "Movies Associated with Leonardo DiCaprio and Christopher Nolan"
  )
```

As we can see, both **Leonardo DiCaprio** and **Christopher Nolan** have a significant number of movies that score highly on our success metric. This confirms their prestige and success in the film industry, as their films consistently receive high ratings and are recognized as successful according to our model.

## Final Validation: Analyzing Bottom Successful Movies

For our final validation, we will select the **bottom 5 successful movies** based on our success metric and manually check if they have won any major awards. This will help us assess whether the model accurately reflects not only commercial success but also critical recognition in the film industry.

```{r}

# Select top 5 movies based on success score
top_movies <- movies |>
  arrange(desc(SuccessNewScore)) |>
  filter(SuccessStatus == "Successful") |>
  slice_tail(n = 5)

summarize <-top_movies |>
  select("Original Name" = primaryTitle, 
         "Average Rating" = averageRating, 
         "Votes" = numVotes, 
         "Success Score" = SuccessNewScore, 
         "Status" = SuccessStatus) |>
  gt() |>
  tab_header(
    title = "Bottom 5 succesful Movies Based on Success Score"
  )
summarize
```

From the bottom 5 successful movies, we can see that all of them have received good recognition. Notably, 3 out of the 5 have won awards, with one being a nominee, which further confirms the accuracy of our model in predicting the success of a movie based on its average rating and number of votes.

# Movie Success Spot Check

1.  **Avatar: The Way of Water (2022)**
    -   **IMDb Rating**: 7.5 \| **Votes**: 512,753
    -   **Awards**: 4 Oscar nominations, **won** Best Visual Effects.
    -   **Box Office**: \$2.3 billion, one of the highest-grossing films ever.
2.  **Wreck-It Ralph (2012)**
    -   **IMDb Rating**: 7.7 \| **Votes**: 467,926
    -   **Awards**: Oscar-nominated for Best Animated Feature; multiple Annie Awards.
    -   **Box Office**: \$471 million globally.
3.  **The Breakfast Club (1985)**
    -   **IMDb Rating**: 7.8 \| **Votes**: 446,348
    -   **Awards**: No major awards, **cult classic**.
    -   **Box Office**: \$51.5 million on a \$1 million budget.
4.  **Independence Day (1996)**
    -   **IMDb Rating**: 7.0 \| **Votes**: 616,858
    -   **Awards**: **Won** 1 Oscar for Visual Effects, nominated for Best Sound.
    -   **Box Office**: \$817 million globally.
5.  **To Kill a Mockingbird (1962)**
    -   **IMDb Rating**: 8.3 \| **Votes**: 337,438
    -   **Awards**: **Won** 3 Oscars, including Best Actor and Best Adapted Screenplay.
    -   **Box Office**: \$13 million, significant for its time.

## Threshold for Success

We examined the "Success Score" metric, which combines both IMDb rating and number of votes. This score reflects both the quality and popularity of a movie. Based on our analysis, we determined that the threshold for a movie to be considered successful is an average rating of **7.9** or higher.

This threshold was derived from the analysis of the top 7 successful movies and the bottom 7 unsuccessful movies in terms of revenue. The score of **7.9** corresponds to the lowest rating of the most successful movies, specifically *Avatar: The Way of Water*.

However, this threshold can be adjusted due to the limited number of successful movies. Adjusting the threshold to **7.5** would represent the average rating of successful films, expanding the criteria from 149 successful movies to 542 movies, thus providing a more inclusive measure of success.

```{r}

library(tidyr)

#changing the threshold to 7.5
TITLE_RATINGS$SuccessStatus <- ifelse(TITLE_RATINGS$SuccessNewScore > 7.5, "Successful", "Unsuccessful")

movies <- TITLE_BASICS |>
  inner_join(TITLE_RATINGS, by = "tconst") |>
  filter(titleType == "movie")

successful_genre <- movies |>
  filter(SuccessStatus == "Successful") |>
  mutate(decade = floor(startYear / 10) * 10) |>
  separate_rows(genres, sep = ",") |>  # Split genres into individual rows
  group_by(decade, genres) |>
  summarise(count = n(), .groups = 'drop')  # Summarise with count

successful_genre_per_decade <- successful_genre |>
  group_by(decade) |>
  filter(count == max(count)) 

ggplot(successful_genre_per_decade, aes(x = decade, y = count, fill = genres)) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Most Successful Genres by Decade",
       x = "Decade",
       y = "Number of Movies",
       fill = "Genre") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1") 

```

For each decade, we observe a variety of genres that have achieved success. However, **Drama** stands out as the most consistently successful genre across all decades. This indicates that drama movies have consistently resonated with audiences and garnered positive reception from critics. While other genres have experienced success in different decades, none have matched the enduring appeal of drama.

## Studying the Evolution of Genres Over Time

We aim to explore which genre consistently yields the most "successes" and identify genres that once produced reliable successes but have since fallen out of favor. To accomplish this, we will exclude drama from our analysis and examine the top 2 genres for each decade, allowing us to highlight emerging trends and shifts in audience preferences.

```{r}
successful_genre_per_decade <- successful_genre |>
  filter(genres != "Drama") |>
  group_by(decade) |>
  arrange(decade, desc(count)) |>  # Arrange by decade and descending count
  slice_head(n = 2)  # Select the top 2 genres per decade

ggplot(successful_genre_per_decade, aes(x = decade, y = count, fill = genres)) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Most Successful Genres by Decade (Excluding Drama)",
       x = "Decade",
       y = "Number of Movies",
       fill = "Genre") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1") 

```

After excluding the drama genre, we observe that **Action** and **Adventure** have been the most successful genres over time, with Action showing significant growth and Adventure demonstrating more consistency. Other genres, such as Mystery, Crime, and Comedy, have had success in the past but lack the consistency seen in Action and Adventure. Additionally, the growth of the **Thriller** genre in the last decade is noteworthy; it has emerged as the third most successful genre following Drama and Action.

## Movies Production vs. Success

We aim to analyze the relationship between the number of movies produced and the number of successful movies. Specifically, we want to determine if the number of successful movies has increased over time and whether the ratio of successful movies to total movies has changed. To achieve this, we will calculate the success ratio by decade and genre, focusing on the top 5 genres: **Drama**, **Action**, **Adventure**, **Thriller**, and **Crime**.

```{r}
#| echo: true
#| message: false
#| warning: false
library(dplyr)
library(gt)

top_genres <- c("Drama", "Action", "Adventure", "Thriller", "Crime")

ratio_successs <- movies |>
  mutate(decade = floor(startYear / 10) * 10) |>
  separate_rows(genres, sep = ",") |>  # Split genres into individual rows if it's a combined column
  filter(genres %in% top_genres) |>  # Focus only on top 5 genre
  group_by(decade, genres) |>
  summarise(
    total_movies = n(), 
    successful_movies = sum(SuccessStatus == "Successful"), 
    .groups = 'drop'
  ) |>
  mutate(successful_ratio = round((successful_movies / total_movies)*100, 2))  # Round to 2 decimals

ggplot(ratio_successs, aes(x = decade, y = successful_ratio, color = genres)) +
  geom_line() +
  geom_point() +
  labs(title = "Percentage of Successful Movies to Total Movies by Decade and Genre",
       x = "Decade",
       y = "Success Ratio",
       color = "Genre") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Set1")
```

With this graph, we can observe the varying ratios of success, indicating the percentage of successful movies within each genre. Overall, there is a clear upward trend in the success of movies across all genres. However, **Adventure** and **Action** have shown the most significant growth since the 1970s, while **Drama** has maintained a consistent performance over time. This highlights the evolving landscape of popular genres and their ability to resonate with audiences across different decades.

```{r}
#| echo: true
#| message: false
#| warning: false
ggplot(ratio_successs, aes(x = decade, y = total_movies, fill = genres)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Number of Movies by Decade and Genre",
       x = "Decade",
       y = "Number of Movies",
       fill = "Genre") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1")

```

From these results, we can clearly see that the number of movies created in the **Drama** genre consistently exceeds that of any other genre by more than double, though the success ratio has increased over time. In contrast, the success ratios for **Adventure** and **Action** have shown growth since the 1970s, while the number of movies produced in these genres has remained steady.

Additionally, we observe that in the last decade, **Thriller** has increased both in the number of movies created and its success ratio, indicating a positive trend for this genre. Overall, there is a decline in movie productions, which can be attributed to the rise of streaming services that have transformed the way movies are produced and consumed. This trend is likely to continue as the industry adapts to new technologies and shifting consumer preferences.

## Hollywood Project

The next step is to create our own movie. To do this, we will first select a genre, two actors, and one director that would be ideal for the film. Here are the criteria we will use:

1.  **Genre**: **Adventure**, combined with some **Action**, as it has proven consistent over time with a high success ratio.
2.  **Actors and Director**: We will select successful individuals from this genre while incorporating some new actors who have found success in the last decade.

```{r}

my_genres <- c("Action", "Adventure")

my_movie <-movies |>
  inner_join(TITLE_PRINCIPALS, by = "tconst") |>
  inner_join(NAME_BASICS, by = "nconst")

get_crew <- my_movie |>
  filter(startYear >= 2008) |>
  separate_rows(genres, sep = ",") |>  
  filter(genres %in% my_genres, SuccessStatus == "Successful")

top_actors <- get_crew |>
  filter(category == "actor" | category == "actress") |>
  group_by(primaryName) |>  
  summarise(successful_movies = n()) |>  # Count the number of successful movies per actor
  arrange(desc(successful_movies)) |>  # Sort by most successful movies
  slice_head(n = 6)  # Select the top 5 actors

top_directors <- get_crew |>
  filter(category == "director") |>  
  group_by(primaryName) |>  
  summarise(successful_movies = n()) |>  
  arrange(desc(successful_movies))|> 
  slice_head(n = 7)  

top_actors |>
  gt() |>
  tab_header(
    title = "Top 5 Actors in Successful Action/Adventure Movies After 2010"
  )

top_directors |>
  gt() |>
  tab_header(
    title = "Most Important Directors in Successful Action/Adventure Movies After 2010"
  )

```

Now that we have identified the actors with the most success in the **Adventure** genre, we will select the top 2 actors and the top director based on their success ratios. To accomplish this, we will calculate the ratio of successful movies to total movies for each actor and director.

Following this analysis, we will create a bar plot to visualize both the total number of movies and the number of successful movies for the selected actors and directors. This will help us clearly see who has performed best in this genre and support our decision-making process for the movie project.

```{r}
#| echo: true
#| message: false
#| warning: false
#parameters to select the actors and directors with our genre. 
selected_actors <- c("Chris Evans", "Hugo Weaving", "Robert Downey Jr.", "Irrfan Khan", "Clark Gregg", "Henry Cavill")
selected_directors <- c("David Yates", "Zack Snyder", "Chris Sanders", "Christopher Nolan", "Dean DeBlois", "Guy Ritchie", "J.J. Abrams")
my_genres <- c("Action", "Adventure")


# Step 2: Filter the data for selected actors and directors
actor_movies <- my_movie |>
  separate_rows(genres, sep = ",") |>  
  filter(genres %in% my_genres & startYear >= 2008 & primaryName %in% selected_actors)|>
  group_by(primaryName) |>
  summarise(total_movies = n())

director_movies <- my_movie |>
  separate_rows(genres, sep = ",") |>  
  filter(genres %in% my_genres & startYear >= 2008 & primaryName %in% selected_directors)|>
  group_by(primaryName) |>
  summarise(total_movies = n())

# we would calculate the success ratio for each actor. 
actor_data <- actor_movies |>
  left_join(top_actors, by = "primaryName") |>
  replace_na(list(successful_movies = 0)) |>
  mutate(success_ratio = successful_movies / total_movies)  # Calculate success ratio

# Reshape data for easier plotting of bars,  this help us to create the plot.
actor_data_long <- actor_data |>
  pivot_longer(cols = c("total_movies", "successful_movies"), 
               names_to = "category", 
               values_to = "count")

# Create the bar plot with a line for the success ratio
ggplot() +
  # Bar plot for total and successful movies
  geom_bar(data = actor_data_long, aes(x = primaryName, y = count, fill = category), 
           stat = "identity", position = "dodge") +
  
  # Line plot for success ratio (on top of the original unreshaped data)
  geom_line(data = actor_data, aes(x = primaryName, y = success_ratio * max(actor_data$total_movies), group = 1), 
            color = "blue", size = 1, linetype = "dashed") +  
  
  # Add points for success ratio
  geom_point(data = actor_data, aes(x = primaryName, y = success_ratio * max(actor_data$total_movies)), 
             color = "blue", size = 3) +
  
  # Labels and titles
  labs(title = "Movies vs Successful Movies (with Success Ratio)",
       x = "Actor",
       y = "Count of Movies",
       fill = "Category") +
  
  # Secondary y-axis for the success ratio
  scale_y_continuous(sec.axis = sec_axis(~./max(actor_data$total_movies), name = "Success Ratio")) +
  
  theme_minimal() +
  theme(legend.position = "bottom")

```

After analyzing the data, we can see that **Clark Gregg** and **Hugo Weaving** have a record of success, each with a **100% success ratio** in **4 movies** after 2008. However, we want to incorporate **Chris Evans** into our movies because, despite his **success ratio being 0.5**, he has appeared in **12 movies**, which is a good indicator of his popularity and appeal to audiences.

```{r}
# Prepare the director data (similar process as for actors)
director_data <- director_movies |>
  left_join(top_directors, by = "primaryName") |>
  replace_na(list(successful_movies = 0)) |>
  mutate(success_ratio = successful_movies / total_movies)  # Calculate success ratio

# Reshape data for easier plotting of bars
director_data_long <- director_data |>
  pivot_longer(cols = c("total_movies", "successful_movies"), 
               names_to = "category", 
               values_to = "count")

# Create the bar plot with a line for the success ratio
ggplot() +
  # Bar plot for total and successful movies
  geom_bar(data = director_data_long, aes(x = primaryName, y = count, fill = category), 
           stat = "identity", position = "dodge") +
  
  # Line plot for success ratio (on top of the original unreshaped data)
  geom_line(data = director_data, aes(x = primaryName, y = success_ratio * max(director_data$total_movies), group = 1), 
            color = "blue", size = 1, linetype = "dashed") +  
  
  # Add points for success ratio
  geom_point(data = director_data, aes(x = primaryName, y = success_ratio * max(director_data$total_movies)), 
             color = "blue", size = 3) +
  
  # Labels and titles
  labs(title = "Movies vs Successful Movies (Directors, After 2010, with Success Ratio)",
       x = "Director",
       y = "Count of Movies",
       fill = "Category") +
  
  # Secondary y-axis for the success ratio
  scale_y_continuous(sec.axis = sec_axis(~./max(director_data$total_movies), name = "Success Ratio")) +
  
  theme_minimal() +
  theme(legend.position = "bottom")




```

Now, we will select our director based on the **success ratio**. In this case, it is an easy pick: we have selected **David Yates**, who has **3 successful movies out of 5** in the genre we are looking for. Although other directors have more movies, the **success ratio** is the most important factor in this case.

## Movie Creation

Our proposed movie is a remake. We analyze films that were successful 25 years ago to identify candidates for a modern adaptation. We select 7 movies within the Action and Adventure genres to maximize the probability of success.

```{r}

library(knitr)

my_genres <- c("Action", "Adventure")

my_movie <-movies |>
  inner_join(TITLE_PRINCIPALS, by = "tconst") |>
  inner_join(NAME_BASICS, by = "nconst") |>
  inner_join(TITLE_CREW, by = "tconst")  # Join with crew data



selection <- my_movie |>
  separate_rows(genres, sep = ",") |>
                
  filter((startYear < 2000) & (SuccessStatus == "Successful") & (genres %in% my_genres)) |>
  group_by(primaryTitle) |>
  filter(n_distinct(genres) == 2) |>  # Keep only movies with both Action and Adventure genres
  arrange(desc(SuccessNewScore)) |>
  distinct(primaryTitle, .keep_all = TRUE)  |>  #Keep only the first occurrence of each movie, then ungroup
  ungroup()   
  
selection |>
  select("Original Name" = primaryTitle, 
         "Average Rating" = averageRating, 
         "Votes" = numVotes,
         "Year" = startYear,
         "Success Score" = SuccessNewScore) |>
  gt() |>
  tab_header(
    title = "Successful Movies with Action and Adventure Genres Before 2000"
  )
```

## Movie Selection: **North by Northwest (1959)**

Based on the table analysis, we have chosen **North by Northwest (1959)** for the following key reasons:

1.  **The movie has not been remade**: This classic suspense thriller remains untouched, providing a unique opportunity for a fresh adaptation. Modern audiences might appreciate a new take on the original storyline.

2.  **Simplified rights acquisition**:

    -   Most of the original **actors and director are no longer living**, which may simplify the process of securing rights.
    -   This could make the adaptation process faster and less complex from a legal standpoint.

A remake of this iconic film can resonate well with contemporary audiences, who may find its themes of suspense, intrigue, and mistaken identity just as compelling today as they were in 1959. A new adaptation, with modern cinematic techniques and storytelling, has the potential to breathe new life into this timeless thriller and mistery.

```{r}

# Extract genres for "North by Northwest"
genres <- TITLE_BASICS |>
  filter(primaryTitle == "North by Northwest") |>
  select(genres)

# Select "North by Northwest" for the remake and display relevant info
north <- my_movie |>
  filter(primaryTitle == "North by Northwest") |>
  select(directors,
         writers,
         primaryName) |>
  distinct()  # Ensure no duplicate rows are included

# Display the selected movie info in a nicely formatted table
north |>
  gt() |>
  tab_header(
    title = "Selected Movie for Remake: North by Northwest (1959)"
  )

```

Based on the Crew group for this movie, the only one alive is Eva Marie Saint which today has 100 years, which would be a good idea to include her in the movie, to give a sense of continuity with the original movie.

## Proposal for Remake of *North by Northwest*

We propose a modern remake of Alfred Hitchcock’s **1959 classic *North by Northwest***, a thrilling action-adventure about mistaken identity and espionage. This project capitalizes on the growing popularity of the action, adventure, and thriller genres, all of which have seen strong growth over the past decade.

**Chris Evans** (success ratio 0.5, 12 films) will play the lead, Roger Thornhill. His global popularity from blockbuster franchises like *Captain America* guarantees mass appeal, particularly among younger audiences. **Hugo Weaving** (100% success ratio, 4 films) will play the sophisticated villain, bringing depth and menace, as he has done in *The Matrix* and *The Lord of the Rings*. **Clark Gregg** (100% success ratio, 4 films) will take on a pivotal supporting role, adding gravitas as a shadowy government figure.

**David Yates**, with a **60% success ratio**, is our chosen director. Known for delivering hits like *Harry Potter* and *Fantastic Beasts*, his ability to handle large-scale action-adventure films with complex narratives ensures this project will be a success.

The consistent rise of the action-adventure genre, paired with the recent resurgence of thrillers, positions *North by Northwest* perfectly for a remake. Our analysis shows that these genres have seen significant growth, and tapping into this trend with a strong cast and proven director will maximize our movie's potential.

We project the film to gross **\$600 to \$800 million globally**, appealing to both classic film fans and modern action audiences.

## Conclusion

This project has provided valuable insights into the film industry, focusing on the success of movies based on IMDb ratings, number of votes, and revenue. We have developed a model to predict the success of a movie based on these factors, and we have validated the model through various spot checks and analyses. We have also explored the evolution of genres over time, identifying trends and patterns that can inform future movie production. also, the creation of our own movie has been a great experience, and we are confident that our proposed remake of *North by Northwest* has the potential to be a success .